---
title: "Assignment1"
output: html_document
date: "2025-10-08"
---
```{r}
# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())
```



## 1. Data preparation
```{r}
# Loading data
df <- read.csv("RealEstate_Georgia.csv", header=T)

# Converting the data column to numeric
df$datePostedString <- as.Date(df$datePostedString, format = "%Y-%m-%d")
df$datePostedString <- as.numeric(format(df$datePostedString, format = "%Y"))

length(df[which(df$datePostedString == 2021),]$datePostedString)
df <- df[which(df$datePostedString == 2021),]

#Check for duplicates
any(duplicated(df$id))
sum(duplicated(df$id))
duplicated_ids <- df$id[duplicated(df$id)]
unique(duplicated_ids)

df[df$id %in% duplicated_ids, ] %>%
  arrange(id)


```
Problem: same information for all variables but different countyID
Since the variable "County" shows the same value for duplicated instances, "CountyId" (5) will be removed due to the inconsistencies. If a numerical representation is needed, we will provide one 

Duplicate removal
```{r}
library(dplyr)

df <- df %>%
  distinct(id, .keep_all = TRUE)
  

any(duplicated(df$id))

```

# 1.1. Variable analysis

Initial analysis
```{r}
summary(df)
```
Converting categorical variables to factors
```{r}
# Converting characters to factor
df[sapply(df, is.character)] <- lapply(df[sapply(df, is.character)], as.factor)

# Converting binary columns (only 0 or 1) to factor
df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))] <- 
  lapply(df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))], as.factor)

summary(df)

```

Variables X.1 (1), X (2), id(3), and time (12) are removed because they don't contribute to predicting the house price. 

Also, stateId (4) is removed because all houses are in Georgia. As well as country (7), datePostedString (8), hasBadGeocode (22), state (16), currency (24).

Variables is_bankOwned (9) and is_forAuction (10) are removed due to having only one observation different from 0. 

Variables livingAreaValue (26) and buildingArea(29) is an exact duplicate of livingArea so it's also removed.

```{r}
summary(df$livingArea == df$livingAreaValue)
summary(df$livingArea == df$buildingArea)
```
# 1.1.1 Target variable : Price
```{r}
summary(df$price)
```
```{r}
hist(df$price, breaks = 30, freq = F, col="cyan")
curve(dnorm(x, mean(df$price), sd(df$price)), add = T)
```
The target variable price does not follow a normal distribution. 


Variable price
--> A lot of outliers (take care of it)

```{r}
v <- df$price

# Summary statistics
summary(v)


# Histogram + Normal curve
hist(v, breaks = 30, freq = FALSE, main = "Histogram of varname", xlab = "varname")
curve(dnorm(x, mean(v, na.rm=TRUE), sd(v, na.rm=TRUE)), add = TRUE, col = "red")
```


```{r}
# Outliers
# Find the threshold

#UPPER LIMITS
varout <- summary(df$price); varout
iqr <- varout[5] - varout[2]; iqr # diff between IQR = Q3 - Q1
umout <- varout[5] + 1.5*iqr; umout # upper limit to identify mild outliers
usout <- varout[5] + 3*iqr; usout # upper limit to identify extreme outliers

#LOWER LIMITS
varout <- summary(df$price); varout

lmout <- varout[5] - 1.5*iqr; umout # upper limit to identify mild outliers
lsout <- varout[5] - 3*iqr; usout # upper limit to identify extreme outliers

boxplot(df$price, horizontal = T, id=list(labels=row.names(df), n=2))

abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers

abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers <- which(df$price < lmout | df$price > umout)
extreme_outliers <- which(df$price < lsout | df$price > usout)

cat("Number of mild outliers:", length(mild_outliers), "\n")
cat("Number of extreme outliers:", length(extreme_outliers), "\n")

```

# 1.1.2 Numerical variables
Variable pricePerSquareFoot


```{r}
v <- df$pricePerSquareFoot

# Summary statistics
summary(v)


# Histogram + Normal curve
hist(v, breaks = 30, freq = FALSE, main = "Histogram of varname", xlab = "varname")
curve(dnorm(x, mean(v, na.rm=TRUE), sd(v, na.rm=TRUE)), add = TRUE, col = "red")
```


```{r}
# Outliers
# Find the threshold

#UPPER LIMITS
varout <- summary(v); varout
iqr <- varout[5] - varout[2]; iqr # diff between IQR = Q3 - Q1
umout <- varout[5] + 1.5*iqr; umout # upper limit to identify mild outliers
usout <- varout[5] + 3*iqr; usout # upper limit to identify extreme outliers

#LOWER LIMITS
varout <- summary(v); varout

lmout <- varout[2] - 1.5*iqr; umout # upper limit to identify mild outliers
lsout <- varout[2] - 3*iqr; usout # upper limit to identify extreme outliers

boxplot(v, horizontal = T, id=list(labels=row.names(df), n=2))

abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers

abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers <- which(v < lmout | v > umout)
extreme_outliers <- which(v < lsout | v > usout)

cat("Number of mild outliers:", length(mild_outliers), "\n")
cat("Number of extreme outliers:", length(extreme_outliers), "\n")

Boxplot(v, id = list(labels = rownames(df), n = 3),
        main = "Boxplot (car::Boxplot) - outliers etiquetados")

```
Has one very extreme outlier: preliminary analysis without it


```{r}

df$pricePerSquareFoot[1282] <- NA
v <- df$pricePerSquareFoot

df$pricePerSquareFoot[df$pricePerSquareFoot == 205000] <- NA
v <- df$pricePerSquareFoot
summary(v)

length(v)


# Summary statistics
summary(v)


# Histogram + Normal curve
hist(v, breaks = 30, freq = FALSE, main = "Histogram of varname", xlab = "varname")
curve(dnorm(x, mean(v, na.rm=TRUE), sd(v, na.rm=TRUE)), add = TRUE, col = "red")
```


```{r}
# Outliers
# Find the threshold

#UPPER LIMITS
varout <- summary(v); varout
iqr <- varout[5] - varout[2]; iqr # diff between IQR = Q3 - Q1
umout <- varout[5] + 1.5*iqr; umout # upper limit to identify mild outliers
usout <- varout[5] + 3*iqr; usout # upper limit to identify extreme outliers

#LOWER LIMITS
varout <- summary(v); varout

lmout <- varout[2] - 1.5*iqr; umout # upper limit to identify mild outliers
lsout <- varout[2] - 3*iqr; usout # upper limit to identify extreme outliers

boxplot(v, horizontal = T, id=list(labels=row.names(df), n=2))

abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers

abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers <- which(v < lmout | v > umout)
extreme_outliers <- which(v < lsout | v > usout)

cat("Number of mild outliers:", length(mild_outliers), "\n")
cat("Number of extreme outliers:", length(extreme_outliers), "\n")

Boxplot(v, id = list(labels = rownames(df), n = 3),
        main = "Boxplot (car::Boxplot) - outliers etiquetados")
```
Observations:
2651: Probable mistake (price ~ 350k/psqft ~  1100) comparing with similar prices
4446, 5080: 2651: Very expensive houses (price and psqft) - maybe not representative?

Variable yearBuilt


```{r}

v <- df$yearBuilt

# Summary statistics
summary(v)


# Histogram + Normal curve
hist(v, breaks = 30, freq = FALSE, main = "Histogram of varname", xlab = "varname")
curve(dnorm(x, mean(v, na.rm=TRUE), sd(v, na.rm=TRUE)), add = TRUE, col = "red")


```
One value (9999) which is NA --> Analysis without it. Cannot likely be imputated


```{r}

df$yearBuilt[df$yearBuilt < 1492 | df$yearBuilt > 2025] <- NA
v <- df$yearBuilt

# Outliers
# Find the threshold

#UPPER LIMITS
varout <- summary(v); varout
iqr <- varout[5] - varout[2]; iqr # diff between IQR = Q3 - Q1
umout <- varout[5] + 1.5*iqr; umout # upper limit to identify mild outliers
usout <- varout[5] + 3*iqr; usout # upper limit to identify extreme outliers

#LOWER LIMITS
varout <- summary(v); varout

lmout <- varout[2] - 1.5*iqr; umout # upper limit to identify mild outliers
lsout <- varout[2] - 3*iqr; usout # upper limit to identify extreme outliers

boxplot(v, horizontal = T, id=list(labels=row.names(df), n=2))

abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers

abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers <- which(v < lmout | v > umout)
extreme_outliers <- which(v < lsout | v > usout)

cat("Number of mild outliers:", length(mild_outliers), "\n")
cat("Number of extreme outliers:", length(extreme_outliers), "\n")

Boxplot(v, id = list(labels = rownames(df), n = 3),
        main = "Boxplot (car::Boxplot) - outliers etiquetados")

```

# Variable longitude
```{r}
v <- df$longitude
summary(v)
```
```{r}
# Missing values
sum(is.na(v))
```

```{r}
# Histogram
hist(v, breaks = 30, freq = F, main="Histogram of Longitude")
curve(dnorm(x, mean(v), sd(v)), add = T)
```
```{r}
# Outliers
# Find the threshold
varout <- summary(v); varout
iqr <- varout[5] - varout[2]; iqr # diff between IQR = Q3 - Q1
umout <- varout[5] + 1.5*iqr; umout # upper limit to identify mild outliers
usout <- varout[5] + 3*iqr; usout # upper limit to identify extreme outliers
lmout <- varout[5] - 1.5*iqr; umout # upper limit to identify mild outliers
lsout <- varout[5] - 3*iqr; usout # upper limit to identify extreme outliers

boxplot(v, horizontal = T, id=list(labels=row.names(df), n=2)) 
abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers
abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers <- which(v < lmout | v > umout)
extreme_outliers <- which(v < lsout | v > usout)

cat("Number of mild outliers:", length(mild_outliers), "\n")
cat("Number of extreme outliers:", length(extreme_outliers), "\n")
```
# Variable latitude
```{r}
v <- df$latitude
summary(v)
```
```{r}
# Missing values
sum(is.na(v))
```

```{r}
# Histogram
hist(v, breaks = 30, freq = F, main="Histogram of Latitude")
curve(dnorm(x, mean(v), sd(v)), add = T)
```
```{r}
# Outliers
# Find the threshold
varout <- summary(v); varout
iqr <- varout[5] - varout[2]; iqr # diff between IQR = Q3 - Q1
umout <- varout[5] + 1.5*iqr; umout # upper limit to identify mild outliers
usout <- varout[5] + 3*iqr; usout # upper limit to identify extreme outliers
lmout <- varout[5] - 1.5*iqr; umout # upper limit to identify mild outliers
lsout <- varout[5] - 3*iqr; usout # upper limit to identify extreme outliers

boxplot(v, horizontal = T, id=list(labels=row.names(df), n=2)) 
abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers
abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers <- which(v < lmout | v > umout)
extreme_outliers <- which(v < lsout | v > usout)

cat("Number of mild outliers:", length(mild_outliers), "\n")
cat("Number of extreme outliers:", length(extreme_outliers), "\n")
```
# Variable livingArea
```{r}
v <- df$livingArea
summary(v)
```
```{r}
# Missing values
sum(is.na(v))
```

```{r}
# Histogram
hist(v, breaks = 30, freq = F, main="Histogram of Latitude")
curve(dnorm(x, mean(v), sd(v)), add = T)
```
```{r}
# Outliers
# Find the threshold
varout <- summary(v); varout
iqr <- varout[5] - varout[2]; iqr # diff between IQR = Q3 - Q1
umout <- varout[5] + 1.5*iqr; umout # upper limit to identify mild outliers
usout <- varout[5] + 3*iqr; usout # upper limit to identify extreme outliers
lmout <- varout[5] - 1.5*iqr; umout # upper limit to identify mild outliers
lsout <- varout[5] - 3*iqr; usout # upper limit to identify extreme outliers

boxplot(v, horizontal = T, id=list(labels=row.names(df), n=2)) 
abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers
abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers <- which(v < lmout | v > umout)
extreme_outliers <- which(v < lsout | v > usout)

cat("Number of mild outliers:", length(mild_outliers), "\n")
cat("Number of extreme outliers:", length(extreme_outliers), "\n")
```

# Variable Bathrooms: num
Bathroom is a discrete numerical variable. 
```{r}
v <- df$bathrooms
summary(v)
```

```{r}
# Missing values
sum(is.na(df$bathrooms))
```
```{r}
# Histogram
hist(df$bathrooms, breaks = 30, freq = F)
curve(dnorm(x, mean(df$bathrooms), sd(df$bathrooms)), add = T)
```
As can be seen in the histogram, some observations have 0 bathrooms. Since a house cannot realistically have 0 bathrooms, these values are replaced with NA to indicate missing data. 

```{r}
# Replace impossible values (0 bathrooms) with NA
v[v == 0] <- NaN
# Missing values
sum(is.na(v))
```
# Variable garageSpaces
garageSpaces is a discrete numerical variable. 
```{r}
v <- df$garageSpaces
summary(v)
```

```{r}
# Missing values
sum(is.na(v))
```
```{r}
# Histogram
hist(v, breaks = 30, freq = F)
curve(dnorm(x, mean(v), sd(v)), add = T)
```


# 1.1.3 Categorical variables


# Variable countyId  -- no posar-la xq al principi hem dit q la treuriem x errors?
```{r}
summary(df$countyId)
```
```{r}
hist(df$countyId, breaks = 30, freq = F)
curve(dnorm(x, mean(df$countyId), sd(df$countyId)), add = T)
```
```{r}
library(dplyr)

correspondance <- df %>%
  group_by(county) %>%
  summarise(n_ids = n_distinct(countyId[countyId != 0]), .groups = "drop") %>%
  filter(n_ids > 1)

# Check if a county has assigned to it more than one countyId
nrow(correspondance)
```



# Variable cityId
```{r}
summary(df$cityId)
```
```{r}
hist(df$cityId, breaks = 30, freq = F)
curve(dnorm(x, mean(df$cityId), sd(df$cityId)), add = T)
```


```{r}
v <- df$cityId

# Frequency table
table(v)

# Barplot
barplot(table(v), main = "Barplot of varname", las = 2, col = "lightblue")

print(sum(df$cityId == 0))

```
Variable cityId has 297 observations of value 0. 

Try to correct 0 codes
```{r}

library(dplyr)

correspondance <- df %>%
  group_by(city) %>%
  summarise(n_ids = n_distinct(cityId[cityId != 0]), .groups = "drop") %>%
  filter(n_ids > 1)

# Check if a city has assigned to it more than one cityId
nrow(correspondance)


```

# Variable event
```{r}

# Frequency table
table(df$event)

# Proportions
prop.table(table(df$event))

# Barplot
barplot(table(df$event), main = "Barplot of event", las = 2, col = "lightblue")


```

# Variable city
```{r}
df$city = factor(df$city)
table(df$city)
barplot(table(df$city), main = "Barplot of city", col="pink")
sum(is.na(df$city))
```

# Variable streetAddress
Every house has a different street Address, we could remove this one.
```{r}
df$streetAddress = factor(df$streetAddress)
table(df$streetAddress)
barplot(table(df$streetAddress), main = "Barplot of streetAddress", col="pink")
sum(is.na(df$streetAddress))
```
# Variable zipcode
```{r}
df$zipcode = factor(df$zipcode)
table(df$zipcode)
barplot(table(df$zipcode), main = "Barplot of zipcode", col="pink")
sum(is.na(df$zipcode))

#Variable hasGarage
```{r}
df$hasGarage = factor(df$hasGarage)
table(df$hasGarage)
barplot(table(df$hasGarage), main = "Barplot of hasGarage", col="pink")
sum(is.na(df$hasGarage))
```

# Variable Bathrooms: factor
```{r}
table(df$bathrooms)
```

```{r}
summary(df$bathrooms)
```
```{r}
df$bathrooms = factor(df$bathrooms)
barplot(table(df$bathrooms), main = "Barplot of Bathrooms", col="orange")
```
As can be seen in the barplot, some observations have 0 bathrooms. Since a house cannot realistically have 0 bathrooms, these values are replaced with NA to indicate missing data. 

```{r}
# Replace impossible values (0 bathrooms) with NA
df$bathrooms[df$bathrooms == 0] <- NaN
barplot(table(df$bathrooms), main = "Barplot of Bathrooms", col="orange")
```


# Variable parking
```{r}
v <- df$parking
table(v)
```

```{r}
summary(v)
```
```{r}
v = factor(v)
barplot(table(v), main = "Barplot of Parking", col="orange")
```

# Variable garageSpaces
```{r}
v <- df$garageSpaces
table(v)
```

```{r}
summary(v)
```
```{r}
v = factor(v)
barplot(table(v), main = "Barplot of garageSpaces", col="orange")
```


#Variable levels
```{r}
df$levels = factor(df$levels)
table(df$levels)
barplot(table(df$levels), main = "Barplot of levels", col="pink")
sum(is.na(df$levels))
```
#Variable levels
```{r}
df$levels = factor(df$levels)
table(df$levels)
barplot(table(df$levels), main = "Barplot of levels", col="pink")
sum(is.na(df$levels))
```

We evaluate the number of unique values in the strings to unify them. 
```{r}
unique(df$levels)
```
```{r}
# Standardize text
df$levels <- tolower(df$levels) #Normalize capitalization and remove extra spaces
df$levels <- trimws(df$levels)

# Group similar values
df$levels[df$levels %in% c(
  "one",
  "one-other",
  "one-manufactured home 1 story",
  "one-mobile home 1 story",
  "one-two story foyer",
  "one-two",
  "one-one and one half",
  "one-one and one half-two"
)] <- "one"

df$levels[df$levels %in% c(
  "one and one half",
  "one and one half-split level",
  "one and one half-multi/split",
  "1.5 story",
  "one and one half-two"
)] <- "one and one half"

df$levels[df$levels %in% c(
  "two",
  "two-split foyer",
  "two-split level",
  "two-two story foyer",
  "two-foyer - 2 story",
  "two-three or more",
  "multi/split-two"
)] <- "two"

df$levels[df$levels %in% c(
  "three or more",
  "three or more-two story foyer",
  "3 story",
  "two-three or more",
  "2.5 story"
)] <- "three or more"

df$levels[df$levels %in% c(
  "multi/split",
  "multi/split-split level",
  "split level",
  "split foyer",
  "tri level",
  "tri-level",
  "tri level-split level",
  "three or more-split level-tri-level"
)] <- "multi/split"

df$levels[df$levels %in% c(
  "manufactured home 1 story",
  "mobile home 1 story"
)] <- "manufactured"

df$levels[df$levels %in% c(
  "0",
  "other-see remarks"
)] <- "other"

# Convert to factor
df$levels <- factor(df$levels,
                    levels = c("one", "one and one half", "two", 
                               "three or more", "multi/split", 
                               "manufactured", "other"))

# Check the result
table(df$levels)
barplot(table(df$levels), main = "Cleaned Levels", col = "lightpink", las = 2)

```


#Variable pool
```{r}
df$pools = factor(df$pool)
table(df$pool)
barplot(table(df$pool), main = "Barplot of pool", col="pink")
sum(is.na(df$pool))
```
#Variable spa
```{r}
df$spa = factor(df$spa)
table(df$spa)
barplot(table(df$spa), main = "Barplot of pool", col="pink")
sum(is.na(df$spa))
```

#Variable isNewConstruction
```{r}
df$isNewConstruction = factor(df$isNewConstruction)
table(df$isNewConstruction)
barplot(table(df$isNewConstruction), main = "Barplot of isNewConstruction", col="pink")
sum(is.na(df$isNewConstruction))
```

#Variable hasPetsAllowed
We see that there's very little houses with pets allowed 51 vs 6041, so we can consider it does not contribute to the model's prediction. We can remove it.
```{r}
df$hasPetsAllowed = factor(df$hasPetsAllowed)
table(df$hasPetsAllowed)
barplot(table(df$hasPetsAllowed), main = "Barplot of hasPetsAllowed", col="pink")
sum(is.na(df$hasPetsAllowed))
```
#Variable homeType
```{r}
df$homeType = factor(df$homeType)
table(df$homeType)
barplot(table(df$homeType), main = "Barplot of homeType", col="pink")
sum(is.na(df$homeType))
```
#Variable county
```{r}
df$county = factor(df$county)
table(df$county)
barplot(table(df$county), main = "Barplot of county", col="pink")
sum(is.na(df$county))
```

## Removal of unnecessary variables

Variables countyId(5) and cityid(6) are also removed because there are inconsistencies such as different codes for the same city, or in 'cityid' codes that are = 0.
Since they represent ids for the categorical values in variables 'county' and 'city', the clean information is already stored in those variables.

```{r}
#Columns removed after first analysis
df <- df[,-c(1:10, 12, 16, 22, 24, 26, 29)]
```

