---
title: "Assignment1"
output:
  pdf_document: default
  html_document: default
date: "2025-10-08"
---
```{r}
# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())
```

# 1. Data preparation
```{r}
# Loading data
df <- read.csv("RealEstate_Georgia.csv", header=T)

# Convert date column to Date format (YYYY-MM-DD) and extract the year as numeric
df$datePostedString <- as.Date(df$datePostedString, format = "%Y-%m-%d")
df$datePostedString <- as.numeric(format(df$datePostedString, format = "%Y"))

# Count rows where year is 2021 and only keep these
length(df[which(df$datePostedString == 2021),]$datePostedString)
df <- df[which(df$datePostedString == 2021),]

# Duplicate quick inspection
dup_rows <- duplicated(df$id) | duplicated(df$id, fromLast = TRUE) # Identify duplicated IDs
dups_df  <- df[dup_rows, ][order(df$id[dup_rows]), ] # Create a df with only duplicated rows, sorted by ID
head(dups_df, 1) # Show first duplicated entry
```
**Problem:** 
1) Some observations have the same information for all variables but different `countyId`. Since the variable `county` shows the same value for duplicated instances, `countyId` (5) will be removed due to the inconsistencies. If a numerical representation is needed, we will provide one. 

```{r}
# Columns to compare (exclude Ids and countyId) without dropping them from df
cols_idx <- setdiff(seq_along(df), c(1:3, 5))

## Comparison by id (which columns differ between duplicates)
by_id <- split(dups_df[, cols_idx, drop = FALSE], dups_df$id)

# Group duplicates by `id` and compare the selected columns
diffs_list <- lapply(by_id, function(sub) {
  difs <- vapply(sub, function(x) length(unique(x)) > 1, logical(1))
  names(difs)[difs]              # solo las columnas que cambian
})


# summary dataframe indicating differences by id
summary_df <- data.frame(
  id       = names(diffs_list),
  n_diff   = lengths(diffs_list),
  diff_vars = ifelse(lengths(diffs_list) == 0, "None",
                     vapply(diffs_list, function(x) paste(x, collapse = ", "), character(1))),
  stringsAsFactors = FALSE
)

# Frequency of how many duplicates have 0, 1, 2, etc. differing variables
table(summary_df$n_diff)

# Frequency of variable combinations that differ (e.g., "countyId", "countyId, price")
sort(table(summary_df$diff_vars), decreasing = TRUE)
```
**Problem:** 
We manage the duplicates separating them in 2 categories.

Case 1) Exact duplicates: duplicates with same information across all variables but different `countyId` (None group).
    - Action: select one of the two duplicates.
    
Case 2) Conflicting duplicates: some duplicates have different information for some variables
    - Action: check these carefully.

```{r}
# Case 1: Exact duplicates
ids_none <- summary_df$id[summary_df$diff_vars == "None"] 
# keep only the 1st occurrence
keep_idx <- ifelse(df$id %in% ids_none, !duplicated(df$id), TRUE) 
df <- df[keep_idx, ]

# check remaining duplicates among "None" group (should be 0)
sum(duplicated(df$id) & df$id %in% ids_none)  
```
```{r}
# Case 2: Duplicates with differences (require inspection)
summary_rest <- summary_df[summary_df$diff_vars != "None", ]

# Frequency of each difference combination (e.g., "price", "time, price")
sort(table(summary_rest$diff_vars), decreasing = TRUE)

## helper: function to see all rows for a given combination
see_comb <- function(combo_str) {
  ids <- summary_rest$id[summary_rest$diff_vars == combo_str]
  out <- df[df$id %in% ids, ]
  out[order(out$id), ]
}
```
The helper function 'see_comb' allows us to watch the duplicates for the different combinations of differences they have (for example, observations with different description, different value for garageSpaces...)

After analyzing all duplicate groups, the conclusions we have reached are as follows:

  * Observations with different description: Since they have the same values for everything else, and the description is not a relevant variable, only one of the two duplicates will be retained.
  
  * Observations with combinations of different time+event, time+price: After checking all observations, it is confirmed that the duplicates correspond to price or event changes. The most recent observations are the ones to retain, since they contain the most recent information.
      ** One particular observation of this case also includes a modification in description, garageSpaces, hasGarage. Since the description hints that the Garage has been renovated to a basement, the most recent observation is also the one to include. 
  
  * One observation with different value for 'garageSpaces': since 'garage' has value 1 for both observations, we conclude the reliable description is the one where 'garageSpaces' is 1, not the one with value 0.
  
  * One observation with different value for 'parking': Since there is no reasonable way to find out which is the correct observation (value 0 or 1), the decision is to remove both duplicates, so as to not work with incorrect information.
  
### Duplicate removal
```{r}
## Identify ids by inconsistency type
ids_desc_only    <- unique(summary_rest$id[summary_rest$diff_vars == "description"])
ids_parking_only <- unique(summary_rest$id[summary_rest$diff_vars == "parking"])
ids_gs_only      <- unique(summary_rest$id[summary_rest$diff_vars == "garageSpaces"])

## IDs with any difference involving 'time' (e.g., time, price, event...) -> keep the row with the SMALLEST time (most recent)
ids_with_time    <- unique(summary_rest$id[grepl("\\btime\\b", summary_rest$diff_vars)])

keep_idx <- rep(TRUE, nrow(df)) # Initialize vector to mark rows to keep

## Rule A: "parking" only -> drop ALL rows for those ids (cannot decide which one is correct)
if (length(ids_parking_only)) {
  keep_idx[df$id %in% ids_parking_only] <- FALSE
}

## Rule B: "description" only -> keep just the first occurrence per id
if (length(ids_desc_only)) {
  w <- df$id %in% ids_desc_only
  keep_idx[w] <- !duplicated(df$id[w])
}

## Rule C: any combo that includes 'time' -> keep the row with the SMALLEST time (most recent)
## - If some times are NA, keep the smallest non-NA; if all are NA, keep the first row.
if (length(ids_with_time)) {
  for (idv in ids_with_time) {
    rows <- which(df$id == idv)
    if (length(rows) >= 2) {
      tvals <- suppressWarnings(as.numeric(df$time[rows]))
      if (all(is.na(tvals))) {
        keep <- rows[1]
      } else {
        # Replace NA with +Inf so they are never the minimum
        t2 <- ifelse(is.na(tvals), Inf, tvals)
        keep <- rows[which.min(t2)]
      }
      keep_idx[rows] <- FALSE
      keep_idx[keep] <- TRUE
    }
  }
}

## Rule D: "garageSpaces" only -> if hasGarage==1 exists, keep the row with garageSpaces==1;
## otherwise keep the first row.
if (length(ids_gs_only)) {
  for (idv in ids_gs_only) {
    rows <- which(df$id == idv)
    if (length(rows) >= 2) {
      gs <- df$garageSpaces[rows]
      hg <- df$hasGarage[rows]
      if (any(hg == 1, na.rm = TRUE) && any(gs == 1 & hg == 1, na.rm = TRUE)) {
        keep <- rows[which((gs == 1) & (hg == 1))[1]]
      } else {
        keep <- rows[1]
      }
      keep_idx[rows] <- FALSE
      keep_idx[keep] <- TRUE
    }
  }
}
df <- df[keep_idx, ]
any(duplicated(df$id))
```

```{r}
# Clean up intermediate data frames
rm(list = setdiff(ls(), "df"))
```

We create a df to store the outliers.

```{r}
# mild and extreme univariate outliers
df$mild_uout <- 0
df$extr_uout <- 0
df$mout <- 0 #multivariate outliers
```

# 2. Data cleaning

## 2.1. Variable analysis

### Initial analysis
```{r}
summary(df)
```
#### Converting categorical variables to factors
```{r}
# Converting characters to factor
df[sapply(df, is.character)] <- lapply(df[sapply(df, is.character)], as.factor)

# Converting binary columns (only 0 or 1) to factor
df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))] <- 
  lapply(df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))], as.factor)
```
Based on both the data summary and the detailed inspection of individual variables, several features will be removed due to limited informational value, redundancy, or detected inconsistencies:

- `X.1`, `X`, `id`, and `time` will be removed as they do not contribute meaningful information for modeling.

- `stateId`, `state`, `country`, `currency`, and `datePostedString` ill be excluded because they show no variability; all records correspond to properties in Georgia (USA), are listed in USD, and fall within the same year.

- `hasBadGeocode` will be removed since it contains the same values, all zero.

- `is_bankOwned` and `is_forAuction` will be dropped due to extremely low variance (each contains only a single non-zero value), providing no practical value and potentially introducing noise.

- `livingAreaValue` and `buildingArea` are exact duplicates of `livingArea`, so they will be removed to avoid redundancy.

- `zipcode` will be excluded because location information is already captured by other variables (for example, city).

- `latitude` and `longitude` will be removed, as geographic coordinates are not relevant for the modeling objectives.

- `streetAddress` will also be removed, given that it uniquely identifies each property without offering generalizable information.

- `cityId` variable, after analyzing it, shows inconsistencies, with several cities associated with multiple different IDs (for example, Atlanta has 21 distinct codes). Since `cityId` is a redundant and unreliable encoding of the city information, it will be removed from the dataset.

- `city` will not be included in the model, as it represents a highly granular categorical variable with many levels, which could lead to overfitting and complicate model interpretation without providing significant predictive value.

- `event` will not be included in the model, as it primarily reflects the listing status rather than intrinsic property features. Moreover, its extreme imbalance (less than 1% of records labeled as “Sold”) limits its usefulness for predictive modeling.


### 2.1.1. Target variable: `Price`
The first variable to be analyzed is Price, as it is a numerical variable representing the target of the analysis. 

The variable contains no missing values, so imputation is not required. Outlier detection identified 561 mild outliers and 32 extreme outliers, on the higher end of the spectrum. 

```{r}
summary(df$price)
sum(is.na(df$price))
```
```{r}
# Histogram + Normal curve
hist(df$price, breaks = 30, freq = F, main="Histogram of Price", col="cyan")
curve(dnorm(x, mean(df$price), sd(df$price)), add = T)
```
As can be seen in the histogram, the target variable `Price` does not follow a normal distribution. The distribution of `Price` is right-skewed, with most values concentrated between approximately 100,000 and 600,000, and a peak around 300,000. This indicates that lower-priced properties are more common, while higher-priced ones are less frequent but still present.

As the price increases, the frequency of observations declines progressively, and only a small number of properties exceed $1,000,000. This long right tail suggests the presence of high-value outliers, typical of luxury properties, we should take this into account as it could affect modeling.

```{r}
ks.test(df$price, "pnorm", mean(df$price), sd(df$price))
```
The Kolmogorov-Smirnov test was performed to confirm whether the `price` variable follows a normal distribution. The test returned a p-value < 2.2e-16, which is statistically significant. Therefore, we reject the null hypothesis of normality and conclude that price does not follow a normal distribution.

We examine the outliers using the Interquartile Range (IQR) method:
```{r}
varout <- summary(df$price); 
iqr <- varout[5] - varout[2]; # diff between IQR = Q3 - Q1
umout <- varout[5] + 1.5*iqr;  # upper limit to identify mild outliers
usout <- varout[5] + 3*iqr;  # upper limit to identify extreme outliers
lmout <- varout[5] - 1.5*iqr;  # upper limit to identify mild outliers
lsout <- varout[5] - 3*iqr;  # upper limit to identify extreme outliers

boxplot(df$price, horizontal = T, id=list(labels=row.names(df), n=2))
abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers
abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers_price <- which(df$price < lmout | df$price > umout)
extreme_outliers_price <- which(df$price < lsout | df$price > usout)

cat("Number of mild outliers:", length(mild_outliers_price), "\n")
cat("Number of extreme outliers:", length(extreme_outliers_price), "\n")
```
The `price` variable shows a substantial number of outliers, with 561 mild and 32 extreme cases, all located in the upper tail. These outliers might reflect the presence of high-value and luxury properties. Since these values carry meaningful market information, they will be retained. 

However, due to their potential influence on model performance, a logarithmic transformation will be considered to reduce skewness and stabilize variance.

### 2.1.2. Numerical variables
#### Variable `pricePerSquareFoot`
```{r}
summary(df$pricePerSquareFoot)
sum(is.na(df$pricePerSquareFoot))

hist(df$pricePerSquareFoot, breaks = 30, freq = FALSE, main = "Histogram of pricePerSquareFoot", xlab = "pricePerSquareFoot")
curve(dnorm(x, mean(df$pricePerSquareFoot, na.rm=TRUE), sd(df$pricePerSquareFoot, na.rm=TRUE)), add = TRUE, col = "red")
```
After examining the distribution of `pricePerSquareFoot`, the maximum value is disproportionately higher than the median, indicating the presence of extreme values or potential outliers. To investigate this further, we apply the Interquartile Range (IQR) method to detect and evaluate these outliers.

```{r}
varout <- summary(df$pricePerSquareFoot); 
iqr <- varout[5] - varout[2];
umout <- varout[5] + 1.5*iqr;
usout <- varout[5] + 3*iqr;
lmout <- varout[2] - 1.5*iqr;  
lsout <- varout[2] - 3*iqr; 

boxplot(df$pricePerSquareFoot, horizontal = T, id = list(labels = rownames(df), n = 1), main = "Boxplot PricePerSquareFoot")
abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers
abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers <- which(df$pricePerSquareFoot < lmout | df$pricePerSquareFoot > umout)
extreme_outliers <- which(df$pricePerSquareFoot < lsout | df$pricePerSquareFoot > usout)

cat("Number of mild outliers:", length(mild_outliers), "\n")
cat("Number of extreme outliers:", length(extreme_outliers), "\n")
```
The variable `pricePerSquareFoot` presents a strong right-skew with a median around 153, while the maximum reaches an unusually extreme value (over 200,000). The IQR analysis detects 258 mild outliers and 51 extreme outliers, confirming the presence of several abnormally high observations. These values likely result from data entry errors or exceptional listings and will require treatment to prevent them from disproportionately influencing the model.

```{r}
df <- df[df$pricePerSquareFoot != 205000, ] # Delete extreme outlier

hist(df$pricePerSquareFoot, breaks = 30, freq = FALSE, main = "Histogram of pricePerSquareFoot", xlab = "pricePerSquareFoot")
curve(dnorm(x, mean(df$pricePerSquareFoot, na.rm=TRUE), sd(df$pricePerSquareFoot, na.rm=TRUE)), add = TRUE, col = "red")
```
After removing the extreme outlier in 205000, the distribution of pricePerSquareFoot remains right-skewed but is now much more realistic and consistent, most observations fall between $100 and $200, with a median of 153 and a maximum around 1,143.

```{r}
varout <- summary(df$pricePerSquareFoot); 
iqr <- varout[5] - varout[2]; 
umout <- varout[5] + 1.5*iqr; 
usout <- varout[5] + 3*iqr; 
lmout <- varout[2] - 1.5*iqr;
lsout <- varout[2] - 3*iqr; 

boxplot(df$pricePerSquareFoot, horizontal = T, id = list(labels = rownames(df), n = 10), main = "Boxplot of PricePerSquareFoot")
abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers
abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers <- which(df$pricePerSquareFoot < lmout | df$pricePerSquareFoot > umout)
extreme_outliers <- which(df$pricePerSquareFoot < lsout | df$pricePerSquareFoot > usout)

cat("Number of mild outliers:", length(mild_outliers), "\n")
cat("Number of extreme outliers:", length(extreme_outliers), "\n")
```

Also, the two most extreme outliers are removed from the dataset, as keeping them could introduce bias and negatively affect predictive model.

```{r}
df <- df[df$pricePerSquareFoot < 1000, ] # Delete 2 extreme outliers

varout <- summary(df$pricePerSquareFoot); 
iqr <- varout[5] - varout[2]; 
umout <- varout[5] + 1.5*iqr; 
usout <- varout[5] + 3*iqr; 
lmout <- varout[2] - 1.5*iqr; 
lsout <- varout[2] - 3*iqr; 

boxplot(df$pricePerSquareFoot, horizontal = T, id = list(labels = rownames(df), n = 10), main = "Boxplot of PricePerSquareFoot")
abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers
abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers_psqft <- which(df$pricePerSquareFoot < lmout | df$pricePerSquareFoot > umout)
extreme_outliers_psqft <- which(df$pricePerSquareFoot < lsout | df$pricePerSquareFoot > usout)

cat("Number of mild outliers:", length(mild_outliers_psqft), "\n")
cat("Number of extreme outliers:", length(extreme_outliers_psqft), "\n")
```

#### Variable `yearBuilt`

```{r}
summary(df$yearBuilt)
sum(is.na(df$yearBuilt))

hist(df$yearBuilt, breaks = 30, freq = FALSE, main = "Histogram of yearBuilt", xlab = "yearBuilt")
curve(dnorm(x, mean(df$yearBuilt, na.rm=TRUE), sd(df$yearBuilt, na.rm=TRUE)), add = TRUE, col = "red")
```

9999 represents a NA, we remove it as the year built likely cannot be imputated.
```{r}
df <- df[df$yearBuilt >= 1700 & df$yearBuilt <= 2025, ] # we only keep in the df the years within a realistic range
hist(df$yearBuilt, breaks = 30, freq = FALSE, main = "Histogram of yearBuilt", xlab = "yearBuilt")
curve(dnorm(x, mean(df$yearBuilt, na.rm=TRUE), sd(df$yearBuilt, na.rm=TRUE)), add = TRUE, col = "red")
```
The distribution of `yearBuilt` shows that most properties were constructed between 1950 and 2010, with a gradual increase in construction frequency over the decades and a peak around the early 2000s. Very few homes were built before 1900, indicating that such early values are rare and may represent historic properties.

```{r}
varout <- summary(df$yearBuilt); 
iqr <- varout[5] - varout[2]; 
umout <- varout[5] + 1.5*iqr; 
usout <- varout[5] + 3*iqr; 
lmout <- varout[2] - 1.5*iqr; 
lsout <- varout[2] - 3*iqr; 

boxplot(df$yearBuilt, horizontal = T, id = list(labels = rownames(df), n = 10), main = "Boxplot yearBuilt")
abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers
abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers_year <- which(df$yearBuilt < lmout | df$yearBuilt > umout)
extreme_outliers_year <- which(df$yearBuilt < lsout | df$yearBuilt > usout)

cat("Number of mild outliers:", length(mild_outliers_year), "\n")
cat("Number of extreme outliers:", length(extreme_outliers_year), "\n")
```
Although the IQR method identifies a set of mild and extreme outliers in `yearBuilt`, early construction years are possible, while very recent years reflect new constructions. Since these observations provide meaningful information about housing age, they are retained rather than removed or imputed.


#### Variable `livingArea`

```{r}
summary(df$livingArea)
sum(is.na(df$livingArea))
```

```{r}
hist(df$livingArea, breaks = 30, freq = F, main="Histogram of livingArea", xlab = "livingArea")
curve(dnorm(x, mean(df$livingArea, na.rm=TRUE), sd(df$livingArea, na.rm=TRUE)), add = TRUE, col = "red")
```
```{r}
varout <- summary(df$livingArea); 
iqr <- varout[5] - varout[2]; 
umout <- varout[5] + 1.5*iqr; 
usout <- varout[5] + 3*iqr; 
lmout <- varout[5] - 1.5*iqr; 
lsout <- varout[5] - 3*iqr; 

boxplot(df$livingArea, horizontal = T, id = list(labels = rownames(df), n = 10), main = "Boxplot of livingArea")
abline(v=umout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=usout,col="red",lwd=2.5) # vertical line for extreme outliers
abline(v=lmout,col="orange",lwd=2) # vertical line for mild outliers
abline(v=lsout,col="red",lwd=2.5) # vertical line for extreme outliers

mild_outliers_area <- which(df$livingArea < lmout | df$livingArea > umout)
extreme_outliers_area <- which(df$livingArea < lsout | df$livingArea > usout)

cat("Number of mild outliers:", length(mild_outliers_area), "\n")
cat("Number of extreme outliers:", length(extreme_outliers_area), "\n")
```
The `livingArea` variable shows a wide range of property sizes, with most homes falling between approximately 1,500 and 3,000 square feet. The IQR method identifies 247 mild outliers and 4 extreme outliers, which correspond to very large homes exceeding 5,000 sq ft. These values are likely not data errors but represent luxury properties with larger living spaces. Therefore, these outliers are retained, as they reflect meaningful variability in property size.


#### Variable `bathrooms`: num
Bathroom is a discrete numerical variable. 
```{r}
summary(df$bathrooms)
sum(is.na(df$bathrooms))
```

```{r}
hist(df$bathrooms, breaks = 30, freq = F, main="Histogram of bathrooms", xlab = "bathrooms")
curve(dnorm(x, mean(df$bathrooms, na.rm=TRUE), sd(df$bathrooms, na.rm=TRUE)), add = TRUE, col = "red")
```
As can be seen in the histogram, some observations have 0 bathrooms. Since a house cannot realistically have 0 bathrooms, these values are replaced with NA to indicate missing data and then later we will impute the variable using the **MICE** algorithm. 

```{r}
df$bathrooms[df$bathrooms == 0] <- NA # Replace impossible values (0 bathrooms) with NA
sum(is.na(df$bathrooms))
```
After inspecting the advertisements for properties with 9 and 10 bathrooms, we confirmed that these values are accurate and correspond to luxury homes. Therefore, they are not treated as outliers, but retained as valid observations reflecting luxury properties in the data set.


#### Variable `bedrooms`: num
Bedrooms is a discrete numerical variable. 
```{r}
summary(df$bedrooms)
sum(is.na(df$bedrooms))
sum(df$bedrooms == 0)
```

```{r}
hist(df$bedrooms, breaks = 30, freq = F, main="Histogram of bedrooms", xlab = "bedrooms")
curve(dnorm(x, mean(df$bedrooms, na.rm=TRUE), sd(df$bedrooms, na.rm=TRUE)), add = TRUE, col = "red")
```

As shown in the histogram, a few observations have 0 bedrooms. After checking the description variable, we found that while a few listings explicitly mention having no bedrooms, most do not provide this information. Therefore, we consider these cases as missing data and replace the 0 values with NA, so that the variable can later be imputed using the **MICE** algorithm.

```{r}
df$bedrooms[df$bedrooms == 0] <- NA # Replace impossible values (0 bedrooms) with NA
sum(is.na(df$bedrooms)) # We check if it was applied correctly
```
 
### 2.1.3. Categorical variables

#### Variable `hasGarage`
```{r}
df$hasGarage = factor(df$hasGarage)
table(df$hasGarage)
prop.table(table(df$hasGarage))
barplot(table(df$hasGarage), main = "Barplot of hasGarage", col="lightblue")
sum(is.na(df$hasGarage))
```
Most properties in the dataset (approximately 63%) have a garage, while the remaining 37% do not. Since garage availability may influence property price and desirability, this variable will be retained.

#### Variable `parking`
```{r}
table(df$parking)
```
```{r}
df$parking = factor(df$parking)
barplot(table(df$parking), main = "Barplot of Parking", col="lightblue")
```
Same as with `hasGarage`, the parking availability could affect the property price and desirability, thus, we maintain the variable.

#### Variable `garageSpaces`
```{r}
table(df$garageSpaces)
```
```{r}
df$garageSpaces = factor(df$garageSpaces)
barplot(table(df$garageSpaces), main = "Barplot of garageSpaces", col="lightblue")
```
Since higher numbers of `garageSpaces` have fewer observations, we consider them as 1 level. 

```{r}
df$garageSpaces <- as.character(df$garageSpaces)
df$garageSpaces[df$garageSpaces>=3] <- "3 or more"
df$garageSpaces <- as.factor(df$garageSpaces)
table(df$garageSpaces)
```
```{r}
barplot(table(df$garageSpaces), main = "Barplot of garageSpaces", col="lightblue")
```

#### Variable `levels`
```{r}
df$levels = factor(df$levels)
table(df$levels)
barplot(table(df$levels), main = "Barplot of levels", col="lightblue")
sum(is.na(df$levels))
```
To simplify the levels variable, some in between levels like the "one and one half" level was reassigned based on its similarity in average price to other groups. After comparing mean prices across levels, it was grouped with "two" to reduce sparsity and improve model interpretability. The same happens with "zero", which was grouped with "one".

```{r}
boxplot(price ~ levels, data = df,
        main = "Price by Level",
        col = "lightblue",
        las = 2)
```

```{r}
# Standardize text formatting (lowercase + trim)
df$levels <- tolower(trimws(df$levels))

# Define categories
one_group <- c(
  "0", "one", "one-other", "one-manufactured home 1 story",
  "one-mobile home 1 story", "one-two story foyer", "one-two",
  "one-one and one half", "one-one and one half-two",
  "manufactured home 1 story", "mobile home 1 story"
)

two_group <- c(
  "one and one half", "one and one half-split level",
  "one and one half-multi/split", "1.5 story", "one and one half-two",
  "two", "two-split foyer", "two-split level", "two-two story foyer",
  "two-foyer - 2 story", "two-three or more", "multi/split-two",
  "two-multi/split"
)

three_group <- c(
  "three or more", "three or more-two story foyer",
  "3 story", "2.5 story", "tri level", "tri-level",
  "tri level-split level", "three or more-split level-tri-level",
  "multi/split", "multi/split-split level",
  "split level", "split foyer"
)
df$levels[df$levels %in% one_group] <- "one"
df$levels[df$levels %in% two_group] <- "two"
df$levels[df$levels %in% three_group] <- "three or more"

# Handle remaining categories (other-see remarks)
df$levels[!df$levels %in% c("one", "two", "three or more")] <- NA

df$levels <- factor(df$levels, levels = c("one", "two", "three or more"))

table(df$levels)
summary(df$levels)
```
```{r}
barplot(table(df$levels), main = "Barplot of levels", col="lightblue")
```

#### Variable `pool`
```{r}
df$pool = factor(df$pool)
table(df$pool)
barplot(table(df$pool), main = "Barplot of pool", col="lightblue")
sum(is.na(df$pool))
```

#### Variable `spa`
```{r}
df$spa = factor(df$spa)
table(df$spa)
barplot(table(df$spa), main = "Barplot of spa", col="lightblue")
sum(is.na(df$spa))
```

#### Variable `isNewConstruction`
```{r}
df$isNewConstruction = factor(df$isNewConstruction)
table(df$isNewConstruction)
barplot(table(df$isNewConstruction), main = "Barplot of isNewConstruction", col="lightblue")
sum(is.na(df$isNewConstruction))
```
The variables `pool`, `spa` and `isNewConstruction` may influence the house price, so we keep them.

#### Variable `hasPetsAllowed`
We see that there's very little houses with pets allowed 51 vs 6041, so we can consider it does not contribute to the model's prediction. We can remove it.
```{r}
df$hasPetsAllowed = factor(df$hasPetsAllowed)
table(df$hasPetsAllowed)
barplot(table(df$hasPetsAllowed), main = "Barplot of hasPetsAllowed", col="lightblue")
sum(is.na(df$hasPetsAllowed))
```

#### Variable `homeType`
```{r}
df$homeType = factor(df$homeType)
table(df$homeType)
barplot(table(df$homeType), main = "Barplot of homeType", col="lightblue")
sum(is.na(df$homeType))
```
Most properties are single-family homes, while townhouses, condos, and especially multi-family units are a minority. This imbalance should be considered in modeling. We can check whether the home type influences the price:

```{r}
boxplot(df$price ~ df$homeType, main="Price by Home Type", col="lightblue")
```
Multi-family units and condos are generally the cheaper, with lower median prices and limited price dispersion. Townhouses show a slightly higher median price compared to single-family homes, but single-family homes have a much wider price range and include numerous luxury properties.

#### Variable `f.price`: cat
We created an additional ordinal variable `f.price` by discretizing the continuous `Price` variable into four quartile-based categories: "Low", "Medium-Low", "Medium-High", and "High". 

```{r}
q <- quantile(df$price, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)

df$f.price <- cut(df$price,
                  breaks = q,
                  include.lowest = TRUE,
                  labels = c("Low", "Medium-Low", "Medium-High", "High"))
table(df$f.price)

barplot(table(df$f.price), main = "Barplot of f.price", col="lightblue")
```

#### Variable `f.yearBuilt`: cat
We created an additional ordinal variable `f.yearBuilt` by discretizing `n.yearBuilt` variable into four new categories: "New", "Recent", "Old" and "Historic". 

```{r}
df$n.yearBuilt <- 2025 - df$yearBuilt
df$f.yearBuilt <- cut(df$n.yearBuilt,
                      breaks = c(-Inf, 10, 30, 100, Inf),
                      labels = c("New", "Recent", "Old", "Historic"),
                      right = TRUE)

table(df$f.yearBuilt)
barplot(table(df$f.yearBuilt), main = "Barplot of f.yearBuilt", col="lightblue")
```

## Removed variable analysis
In addition to the variables previously discussed, `countyId` and `cityId` will also be removed due to inconsistencies, such as multiple codes representing the same city, and `cityId` values equal to 0. 

As we said before, these variables are redundant, as they simply encode the categorical information already present in `county` and `city`, which are cleaner and more interpretable. 

Variables `hasPetsAllowed` and `event` will be removed because of the big class imbalance.

```{r}
df2 <- df[,-c(1:12, 15:24, 26, 29, 37, 39)]
```

Finally, all the variables removed from the dataset were excluded due to low informational value, redundancy, or lack of variability. The following variables were discarded:

- X.1 (1)
- X (2)
- id (3)
- stateId (4)
- countyId (5)
- cityId 
- country (7) 
- datePostedString (8)
- is_bankOwned (9)
- is_forAuction (10)
- event (11)
- time (12)
- city (15)
- state (16)
- yearBuilt (17)
- streetAddress (18)
- zipcode (19)
- longitude (20)
- latitude (21)
- hasBadGeocode (22)
- description (23) 
- currency (24)
- livingAreaValue (26)
- buildingArea (29)
- hasPetsAllowed (37)
- county (39)

The dataset obtained after the preprocessing contains the following variables:
- 6 numerical variables: `price`, `pricePerSquareFoot`, `n.yearBuilt`, `livingArea`, `bathrooms`, `bedrooms`.
- 10 categorical variables: `parking`, `garageSpaces`, `hasGarage`, `levels`, `pool`, `spa`, `isNewConstruction`, `homeType`, `f.price`, `f.yearBuilt`.

```{r}
dfnum <- df2[,c(1:5, 18)] # Numerical variables
dfcat <- df2[,c(6:13, 17, 19)] # Categorical variables
dfout <- df2[,c(14:16)]
```
```{r}
summary(dfcat)
```

To explore potential linear dependencies among the numerical features, we computed their correlation matrix.

```{r}
require(corrplot)
par(mfrow=c(1,1)) 
corrplot(cor(dfnum, use = "complete.obs"), method = "number")
```

The correlation matrix shows that `price` is strongly associated with `livingArea` and `bathrooms`, while features such as `yearBuilt` and `garageSpaces` exhibit only weak correlations with the target variable. Additionally, some predictors (e.g., `livingArea`, `bathrooms`, and `bedrooms`) are highly correlated with each other, which may indicate potential multicollinearity.


## 2.2. Missing values
In the dataset there are missing values in variables: bedrooms, bathrooms and levels.

```{r}
summary(df2)
colSums(is.na(df2))
na_per_row <- rowSums(is.na(df2))
table(na_per_row)

df2[rowSums(is.na(df2)) > 1, ]
```
The problem is that in 11 observations, there is a missing value for both bedrooms and bathrooms. We will impute both variables. 

```{r}
library(mice)    

vars <- c(
  "bathrooms", "bedrooms", "levels",
  "livingArea", "price", "pricePerSquareFoot",
  "garageSpaces", "hasGarage", "parking",
  "pool", "spa",
  "homeType", "isNewConstruction",
  "n.yearBuilt"
)

df_mice <- df2[vars]

ini  <- mice(df_mice, maxit = 0, printFlag = FALSE)
meth <- ini$method

meth["bathrooms"] <- "pmm"
meth["bedrooms"]  <- "pmm"
meth["levels"]    <- "polyreg"

imp <- mice(
  df_mice,
  m = 5,
  maxit = 5,
  method = meth,
  seed = 123)

completed_df <- complete(imp)
summary(completed_df)

vars_no_impute <- setdiff(names(df2), vars)
df3 <- cbind(completed_df, df2[vars_no_impute])
```
To validate the results of the imputation, we compare the quantiles before and after to see if they have changed. They have not changed.
```{r}
quantile(df$bathrooms, probs=seq(0,1,by=0.1), na.rm=T)
quantile(df3$bathrooms, probs=seq(0,1,by=0.1), na.rm=F)
quantile(df$bedrooms, probs=seq(0,1,by=0.1), na.rm=T)
quantile(df3$bedrooms, probs=seq(0,1,by=0.1), na.rm=F)
```

## 2.3. Detection of outliers
### 2.3.1. Univariate outliers
We have managed some of the outliers in the previous section, the ones we have not removed are taken into account here:
```{r}
dfout$extr_uout <- as.numeric(as.character(dfout$extr_uout))
dfout$mild_uout <- as.numeric(as.character(dfout$mild_uout))

# all rows that have an extreme outlier for some variable
total_extreme_outliers <- unique(c(
  extreme_outliers_price,
  extreme_outliers_psqft,
  extreme_outliers_year,
  extreme_outliers_area
))

dfout$extr_uout[total_extreme_outliers] <-1

total_mild_outliers <- unique(c(
  mild_outliers_price,
  mild_outliers_psqft,
  mild_outliers_year,
  mild_outliers_area
))

dfout$mild_uout[total_mild_outliers] <-1
```

### 2.3.2. Multivariate outliers
To identify observations that behave unusually when considering all numerical variables together, we applied a multivariate outlier analysis using Mahalanobis distance. This method detects data points that stand out from the general pattern, even if they do not appear extreme in individual variables. The flagged observations were then further inspected as potential multivariate outliers.

```{r}
library(chemometrics) 
dfnum <- df3[,c(1:2, 4:6, 14)]

res.mout <- Moutlier(dfnum,quantile=0.999, plot=F) 
str(res.mout) # list of: 1. md: Mahalanobis distance / 2. rd / 3. cutoff (threshold)

par(mfrow=c(1,1))
plot(res.mout$md,res.mout$rd,col="cyan",pch=19)
res.mout$cutoff
abline(h=res.mout$cutoff,col="red",lwd=2)
abline(v=res.mout$cutoff,col="red",lwd=2)
text(res.mout$md,res.mout$rd,label=row.names(df),cex=0.5) # we can see that physicians and general man are the identified outliers (>cutoff)

ll <- which((res.mout$md>res.mout$cutoff) & (res.mout$rd>res.mout$cutoff)); ll # observations with md and rd > cutoff
length(ll)
```
```{r}
dfout$mout <- as.numeric(as.character(dfout$mout))
dfout$mout[ll] <-1
df3$mout <- dfout$mout
df3 <- df3[!(rownames(df3) %in% c('1715','4157')),]
```

# 3. Address tests to discard serial correlation
```{r}
acf(df3$price) 
```

The ACF plot of the variable price shows significant positive autocorrelation at the first lags, indicating that nearby observations are not independent. Therefore, serial correlation cannot be discarded, and the temporal structure of the data should be taken into account in the modelling process.

Since the temporal order of the observations is not relevant for the predictive model and autocorrelation was observed in the target variable, the dataset was randomly shuffled to remove artificial serial dependence.

```{r}
ll <- sample(1:nrow(df3), nrow(df3))
df4 <- df3[ll, ]
acf(df4$price) 
```

# 4. Profiling
Profiling for the target variable price will be done using the FactoMineR package.
  - As a numerical variable `price`: using 'condes' method. 
  - As a factor `f.price`: using 'catdes' method. 

The factor will help with interpretation, so it will be easier to see discrepancies between groups.
```{r}
# Detect and clean variables
one_level_vars <- names(df3)[sapply(df3, function(x) is.factor(x) && nlevels(x) < 2)] # Categorical variables with fewer than 2 levels
constant_vars <- names(df3)[sapply(df3, function(x) length(unique(x[!is.na(x)])) < 2)] # Identify columns that are all NA or constant numeric values

vars_to_remove <- unique(c(one_level_vars, constant_vars))

df_prof <- df3[, !(names(df3) %in% vars_to_remove)]
```

```{r}
# Analysis with price as numerical
library(FactoMineR)
res.con <- condes(df_prof, which(names(df_prof) == 'price'))

res.con$quanti
res.con$quali
res.con$category
```

## 4.1 Price as numeric

### Numeric Predictors (`res.con$quanti`)
- Strongest positive associations:  
  `livingArea` (~0.72), `bathrooms` (~0.68), `pricePerSquareFoot` (~0.51), `bedrooms` (~0.46).  
  → Larger homes with more bathrooms and bedrooms tend to have higher prices. 
  
- Moderate association:  
  `yearBuilt` (~0.24) shows newer constructions are slightly more expensive.  

### Categorical Predictors (`res.con$quali` and `$category`)
- Most explanatory factors (by R²):  
  `f.price` (as expected), `levels`, `garageSpaces`, `hasGarage`, `spa`, `f.yearBuilt`, `parking`, `pool`.  
- Categories associated with above-average price:  
  `f.price=High`, `garageSpaces=3 or more`, `levels=two` / `three or more`,  
  `hasGarage=1`, `pool=1`, `spa=1`, `f.yearBuilt=New/Recent`,  
  `parking=1`, `homeType=SINGLE_FAMILY`, `event_status=Active`. 
  
- Categories associated with below-average price:  
  `levels=one`, `garageSpaces=0–2`, `hasGarage=0`, `pool=0`, `spa=0`,  
  `f.yearBuilt=Old/Historic`, `homeType=CONDO` or `MULTI_FAMILY`, `parking=0`, `event_status=Sold`.
- Interpretation:  
  More stories, larger garages, newer builds, and amenities (pool/spa/parking) mark higher-value properties;  
  condos, no garage, and older homes pull prices down.

```{r}
# Analysis with price as a factor
res.cat <- catdes(df_prof, which(names(df_prof) == 'f.price'))
res.cat$quanti.var
res.cat$quanti
res.cat$test.chi2
res.cat$category
```

## 4.2. Price as Factor 

### Numeric Predictors (`res.cat$quanti.var` and `$quanti`)
- Top numeric discriminators (by Eta²):  
  `price` (trivial, derived), `livingArea`, `bathrooms`, `pricePerSquareFoot`, `bedrooms`, `yearBuilt`. 
- Profiles by price band:
  - Low: small living area, fewer bathrooms/bedrooms, low price per sq. ft, older construction.  
  - Medium-Low: slightly newer but still modest size and rooms.  
  - Medium-High: newer homes, larger, more rooms, higher price per sq. ft.  
  - High: very large, many rooms/baths, newer builds, highest price per sq. ft.  
  
- Interpretation:  
  Size, number of rooms, and build year cleanly separate price bands.

### Categorical Predictors (`res.cat$test.chi2` and `$category`)
- Most significant categorical links:  
  `hasGarage`, `garageSpaces`, `levels`, `f.yearBuilt`, `parking`, `spa`, `pool`, `homeType`. 
  
- Patterns by band:
  - Low: overrepresented by `no garage`, `garageSpaces=0`, `levels=one`, `parking=0`, `old/historic`,  
    `no spa/pool`, `MULTI_FAMILY` or `CONDO`, and `Sold` listings.  
  - Medium-Low: moderate garage/parking, mostly `spa=0`, some condos, underrepresented in high amenities.    - Medium-High: frequent `garageSpaces=2`, `levels=two`, `parking=1`, `new/recent`, `SINGLE_FAMILY`.  
  - High: dominant in `garageSpaces>=3`, `spa=1`, `pool=1`, `levels>=2`, `newer`, `parking=1`,  
    and `SINGLE_FAMILY`; underrepresented in `CONDO`, `MULTI_FAMILY`, or `old` stock.  
    
Note:
`f.price` and `pricePerSquareFoot` are derived from `price`.  
They are valid for descriptive profiling but should  be used as independent predictors  
in any model aiming to explain or forecast `price`.

# 5. Export to clean csv
```{r}
write.csv(df4, "House_Clean.csv", row.names = FALSE)
```