---
title: "Model_SIM.Rmd"
output:
  pdf_document: default
  html_document: default
date: "2025-10-24"
---
```{r}
# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())

library(car)
library(FactoMineR)
library(lmtest)
library(effects)
library(AER)
library(MASS)
```

```{r}
df <- read.csv("House_Clean2.csv", header = TRUE)

df[sapply(df, is.character)] <- lapply(df[sapply(df, is.character)], as.factor)

# Converting binary columns (only 0 or 1) to factor
df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))] <- 
  lapply(df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))], as.factor)

summary(df)
```

```{r}
num_vars <- names(df)[sapply(df, is.numeric)]
cor(df[,num_vars]) 
plot(df[,num_vars])
```
## Modeling using explanatory numeric variables

How to define the best model? Iterative process.
Suggestion: after data preparation etc. 
    Modelling
  1) Use numeric variables.
Try to find the best transformation needed to the target variable or the explanatory variables (improve the linear )
  2) Add factors (only the main effects)
  3) Add interactions
  
MLR has explanatory variables with only numeric 
When I add factors it becomes GLM 

# Model: numeric variables
We build a model with all numeric variables. As livingArea and pricePerSquareFoot perfectly describe the target, we use livingArea because it's more correlated with the target price.

Analyzing the p-values for the first model (with only numeric variables), we observe that n.yearBuilt has a p-value > 0.05 so we remove this from the model.

Analyzing the variance inflation factor (VIF), we observe that all values are below 5, indicating that the predictors are not highly linearly correlated, so multicollinearity is not a problem.

```{r}
head(df)
m0 <- lm(price ~bedrooms+bathrooms+livingArea+n.yearBuilt, data=df[,num_vars])
summary(m0)
vif(m0)
```
We use boxcox transformation and we observe that lambda is near 0.5, so we apply sqrt transformation to our target price.

We compare two models m1 and m2, which are the models with sqrt of price and the other variables, with and without n.yearsBuilt. In the anova test, p-value < 0.05, so we reject the null hypothesis, meaning that the big model is better, so we keep n.yearBuilt. We also check the AIC value, and this is confirmed because AIC for the model 2 is bigger than AIC for the model 1.

bedrooms lambda negatiu, pr < 0.05 -> reject H0 -> relació no es lineal -> transformació -> no transformació
bathrooms lambda positiu, pr < 0.05 -> reject H0 -> relació no es lineal -> transformació -> sqrt
livingArea lambda positiu, pr < 0.05 -> reject H0 -> relació no es lineal -> transformació -> sqrt


```{r}
boxcox(price ~bedrooms+bathrooms+livingArea, data=df[,num_vars]) #we see that sqrt is the suitable tranformation
boxTidwell(sqrt(price) ~ bedrooms+bathrooms+livingArea+n.yearBuilt, data=df[,num_vars])

#we use sqrt
m1 <- lm(sqrt(price) ~bedrooms+bathrooms+livingArea+n.yearBuilt, data=df[,num_vars])
summary(m1)
vif(m1)
step(m1)
AIC(m1)

#we remove n.yearBuilt
m2 <- lm(sqrt(price) ~ bedrooms+bathrooms+livingArea, data=df[,num_vars])
summary(m2)
vif(m2)
anova(m2,m1)
step(m2)
AIC(m2)

#we use the ^1/3 transformation
m3 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt, data=df[,num_vars])
summary(m3)
vif(m3)
anova(m3,m1)
step(m3)
AIC(m3)
```
We keep m3 as the numeric model.

```{r}
marginalModelPlots(m3, id=list(n=0))
residualPlots(m3, id=list(n=0))
avPlots(m3, id=list(n=0))
crPlots(m3, id=list(n=0))
influencePlot(m3, id=list(n=3))
plot(m3)
plot(allEffects(m3))
```
### Influent data assesment
We see a priori influent data in our dataset, so before moving on to the next step, we remove them.

```{r}
llres<-which(abs(rstudent(m3))>3); llres
llhat<-which(hatvalues(m3)>3*length(coef(m3))/nrow(df));llhat
llcoo<-Boxplot(cooks.distance(m3), id=list(n=10,labels=row.names(df)))
influencePlot(m3)
df[llres,]
ll<-which(row.names(df) %in% c("611", "1590","5415"));ll

df<-df[-ll,]
```
```{r}
# same model without suspicious influent data
m3_1 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt, data=df[,num_vars])
summary(m3_1)
vif(m3_1)
#anova(m3_1,m3)
step(m3_1)
AIC(m3_1)

marginalModelPlots(m3_1, id=list(n=0))
residualPlots(m3_1, id=list(n=0))
avPlots(m3_1, id=list(n=0))
crPlots(m3_1, id=list(n=0))
influencePlot(m3_1, id=list(n=3))
plot(m3_1)
plot(allEffects(m3_1))

```


# Model: numeric variables + factors
After applying condes() to se which factors are more related to Price.

Levels: checking the level factor we observe that p-value < 0.01, we reject the H0 that the shorter model is better, so we keep the model with levels.

```{r}
m4 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt+levels, data=df)
summary(m4)
vif(m4)
AIC(m4)
step(m4)
anova(m3_1,m4)

marginalModelPlots(m4, id=list(n=0))
residualPlots(m4, id=list(n=0))
avPlots(m4, id=list(n=0))
crPlots(m4, id=list(n=0))
influencePlot(m4, id=list(n=3))
plot(m4)
plot(allEffects(m4))
```
```{r}
m5 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt+levels+garageSpaces, data=df)
summary(m5)
vif(m5)
AIC(m5)
step(m5)
anova(m4,m5)

marginalModelPlots(m5, id=list(n=0))
residualPlots(m5, id=list(n=0))
avPlots(m5, id=list(n=0))
crPlots(m5, id=list(n=0))
influencePlot(m5, id=list(n=3))
plot(m5)
plot(allEffects(m5))
```

```{r}
m5 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt+levels+garageSpaces, data=df)
summary(m5)
vif(m5)
AIC(m5)
step(m5)
anova(m5,m4)

marginalModelPlots(m5, id=list(n=0))
residualPlots(m5, id=list(n=0))
avPlots(m5, id=list(n=0))
crPlots(m5, id=list(n=0))
influencePlot(m5, id=list(n=3))
plot(m5)
```

```{r}
m6 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt+levels+garageSpaces+hasGarage, data=df)
summary(m6)
vif(m6)
AIC(m6)
step(m6)
anova(m5,m6)

marginalModelPlots(m6, id=list(n=0))
residualPlots(m6, id=list(n=0))
avPlots(m6, id=list(n=0))
crPlots(m6, id=list(n=0))
influencePlot(m6, id=list(n=3))
plot(m6)
```
```{r}
m7 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt+levels+garageSpaces+hasGarage+spa, data=df)
summary(m7)
vif(m7)
AIC(m7)
step(m7)
anova(m7,m6)

marginalModelPlots(m7, id=list(n=0))
residualPlots(m7, id=list(n=0))
avPlots(m7, id=list(n=0))
crPlots(m7, id=list(n=0))
influencePlot(m7, id=list(n=3))
plot(m7)
```
```{r}
m8 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt+levels+garageSpaces+hasGarage+spa+parking, data=df)
summary(m8)
vif(m8)
AIC(m8)
step(m8)
anova(m8,m7)

marginalModelPlots(m8, id=list(n=0))
residualPlots(m8, id=list(n=0))
avPlots(m8, id=list(n=0))
crPlots(m8, id=list(n=0))
influencePlot(m8, id=list(n=3))
plot(m8)
```

```{r}
m9 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt+levels+garageSpaces+hasGarage+spa+parking+pool, data=df)
summary(m9)
vif(m9)
AIC(m9)
step(m9)
anova(m9,m8)

marginalModelPlots(m9, id=list(n=0))
residualPlots(m9, id=list(n=0))
avPlots(m9, id=list(n=0))
crPlots(m9, id=list(n=0))
influencePlot(m9, id=list(n=3))
plot(m9)
```

```{r}
m10 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt+levels+garageSpaces+hasGarage+spa+parking+pool+homeType, data=df)
summary(m10)
vif(m10)
AIC(m10)
step(m10)
anova(m10,m9)

m10_1 <- lm(sqrt(price) ~bedrooms+bathrooms+I(livingArea^(1/3))+n.yearBuilt+levels+garageSpaces+hasGarage+spa+parking+homeType, data=df)
summary(m10_1)
vif(m10_1)
AIC(m10_1)
step(m10_1)
anova(m10_1,m10)

#marginalModelPlots(m10, id=list(n=0))
#residualPlots(m10, id=list(n=0))
#avPlots(m10, id=list(n=0))
#crPlots(m10, id=list(n=0))
#influencePlot(m10, id=list(n=3))
#plot(m10)
```
```{r}
#amb sqrt
m19 <- lm(sqrt(price) ~bedrooms+bathrooms+sqrt(livingArea)+n.yearBuilt, data=df)
summary(m19)
vif(m19)
AIC(m19)
step(m19)
#anova(m19,m9)

m20 <- lm(sqrt(price) ~bedrooms+bathrooms+sqrt(livingArea)+n.yearBuilt+levels+garageSpaces+hasGarage+spa+parking+pool+homeType, data=df)
summary(m20)
vif(m20)
AIC(m20)
step(m20)
anova(m19,m20)

```

