---
title: "Assignment 1: Model"
author: 'Laia Jané, Elisa Müller, Runxiao Qiu'
output:
  pdf_document: default
  html_document: default
date: "2025-11-09"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message    = FALSE,   # oculta mensajes
  warning    = FALSE,   # oculta warnings
  fig.align  = "center",
  fig.width  = 5,       # tamaño base de los gráficos (en pulgadas)
  fig.height = 3.5,
  dpi        = 120,
  out.width  = "70%"   # reduce tamaño en la página
)
```

```{r, include=FALSE}
if(!is.null(dev.list())) dev.off()
rm(list=ls())

library(car)
library(FactoMineR)
library(lmtest)
library(effects)
library(AER)
library(MASS)
```

# 6. Build the model

```{r, include=FALSE}
df <- read.csv("House_Clean.csv", header = TRUE)

df[sapply(df, is.character)] <- lapply(df[sapply(df, is.character)], as.factor)

# Converting binary columnsto factor
df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))] <- 
  lapply(df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))], as.factor)
```


```{r, include=FALSE}
summary(df)
```

```{r}
num_vars <- names(df)[sapply(df, is.numeric)]
cor(df[,num_vars]) 
```

## 6.1. Modeling using explanatory numeric variables

Once the data were prepared and cleaned, the following steps were carried out to define the **best model**:

1)  Start with the **numeric variables** and check whether any transformations are needed on the target or explanatory variables to improve the linear relationship.
2)  Add **categorical factors**, including only their main effects, to see how they contribute to the model.
3)  Include **interactions** between relevant variables to capture combined effects that might improve model performance.

### 1) Model: numeric variables

We start building the model considering all numeric variables except `pricePerSquareFoot`. As `livingArea` and `pricePerSquareFoot` perfectly describe the target, between the two, `livingArea` was selected because its correlation with the target variable is higher (0.72) than that of `pricePerSquareFoot`. Moreover, using `pricePerSquareFoot` would be conceptually inconsistent, since it is mathematically derived from the target variable. Including both variables would also introduce severe multicollinearity, as they are functionally dependent.

To confirm this, two models were created: one including all numeric variables and another excluding `pricePerSquareFoot`. The second model was retained, as the $R^2$ value increased drastically when both variables were included, which is a clear indicator of multicollinearity.

```{r}
# yearBuilt as numerical variable
m0 <- lm(price ~ bedrooms+bathrooms+livingArea+n.yearBuilt, data=df)
summary(m0)
vif(m0); AIC(m0)
```

```{r}
# yearBuilt as factor
m0_f <- lm(price ~ bedrooms+bathrooms+livingArea+f.yearBuilt, data=df)
summary(m0_f)
vif(m0_f); AIC(m0_f)
```

From the two models testes, the model with `f.yearBuilt` performed slightly better; it had a higher $R^2$ (0.56 vs. 0.55), a lower AIC, and a smaller residual error. All variables are statistically significant, and the VIF values (all below 5) confirm that there are no multicollinearity issues. Therefore, the model with `f.yearBuilt` was selected.

#### Model transformations

After fitting the initial models, a Box–Cox transformation suggested that a **square root transformation** of the target variable was the most suitable. A Box–Tidwell test was then applied to check the linearity between the transformed target and the predictors. The test indicated that `livingArea` showed a non-linear relationship with `price`, so a square root transformation was also applied to this variable.

Three models were then compared:

-   m1: sqrt(price) \~ bedrooms + bathrooms + livingArea + f.yearBuilt
-   m2: sqrt(price) \~ bedrooms + bathrooms + livingArea
-   m3: sqrt(price) \~ bedrooms + bathrooms + sqrt(livingArea) + f.yearBuilt

Model m3 achieved the lowest AIC (52013), the highest R² (0.575), and all VIFs below 5, indicating no multicollinearity. Therefore, m3 was selected.

```{r}
boxcox(price ~ bedrooms+bathrooms+livingArea+n.yearBuilt, data=df) #we see that sqrt is the suitable tranformation
boxTidwell(sqrt(price) ~ bedrooms+bathrooms+livingArea+n.yearBuilt, data=df) #we put sqrt to livingArea

# Model: srt(price)
m1 <- lm(sqrt(price) ~ bedrooms+bathrooms+livingArea+f.yearBuilt, data=df)
#summary(m1)
#vif(m1); step(m1); AIC(m1)

# Model: srt(price) and without f.yearBuilt 
m2 <- lm(sqrt(price) ~ bedrooms+bathrooms+livingArea, data=df)
#summary(m2)
#vif(m2); step(m2); AIC(m2)
#anova(m2,m1)

# Model: srt(price) and sqrt(livingArea)
m3 <- lm(sqrt(price) ~ bedrooms+bathrooms+sqrt(livingArea)+f.yearBuilt, data=df)
summary(m3)
# step(m3)
vif(m3); AIC(m3)
anova(m3,m1)
```

#### Plots of m3 model

The marginal and residual plots show an approximately linear relationship between the predictors and the transformed response. The residual plots display a random scatter around zero with no clear pattern, indicating homoscedasticity and a good model fit.

The Q–Q plot shows a deviation in the tails, but the residuals are approximately normally distributed. The Scale–Location and Residuals vs Fitted plots suggest constant variance across fitted values.

From the Cook’s distance and Residuals vs Leverage plots, a few influential points are detected.

```{r}
# Plots model with transformations
marginalModelPlots(m3, id=list(n=0))
residualPlots(m3, id=list(n=0))
avPlots(m3, id=list(n=0))
crPlots(m3, id=list(n=0))
# influencePlot(m3, id=list(n=3))
par(mfrow=c(2,2))
plot(m3)
par(mfrow=c(1,1))
plot(allEffects(m3))
```

#### Model with bathrooms and bedrooms as factors

We also tested whether treating bedrooms and bathrooms as categorical variables would improve the model. However, the version using them as numeric variables performed better, with a higher $R^2$ (0.5725 vs. 0.5542) and a lower AIC (67529 vs. 67762). Therefore, we kept both variables as numeric.

### 2) Model: all factors

```{r}
# Model with all factors
m_full <- lm(sqrt(price) ~ bedrooms+bathrooms+sqrt(livingArea)+f.yearBuilt+levels+garageSpaces+hasGarage+spa+parking+pool+homeType, data=df)

# summary(m_full)
Anova(m_full)
```

We see that p-values for hasGarage and spa are lower, we try to exclude them from the model

```{r}
# Model without hasGarage
m_full2 <- lm(sqrt(price) ~ bedrooms+bathrooms+sqrt(livingArea)+f.yearBuilt+levels+garageSpaces+spa+parking+pool+homeType, data=df)
#summary(m_full2); Anova(m_full2)

# Model without hasGarage and spa
m_full3 <- lm(sqrt(price) ~ bedrooms+bathrooms+sqrt(livingArea)+f.yearBuilt+levels+garageSpaces+parking+pool+homeType, data=df)
#summary(m_full3); Anova(m_full3)

# Model without spa
m_full4 <- lm(sqrt(price) ~ bedrooms+bathrooms+sqrt(livingArea)+f.yearBuilt+levels+garageSpaces+hasGarage+parking+pool+homeType, data=df)
#summary(m_full4); Anova(m_full4)

m_step <- step(m_full3, direction = "both", trace = TRUE)
summary(m_step)
AIC(m_step)
```

After adding the categorical variables related to the house features, several models were compared. All of them showed similar performance, with $R^2$ around 0.59. The simplest model (m_full3), which excludes hasGarage and spa, was selected because it achieves almost the same explanatory power (Adjusted $R^2=0.594$) as the more complex versions, while keeping a lower AIC (67513) and fewer parameters.

All predictors in the final model are statistically significant (p < 0.05), and the VIF values remain below 5, confirming that there is no multicollinearity.

```{r}
marginalModelPlots(m_full3, id=list(n=0))
residualPlots(m_full3, id=list(n=0))
avPlots(m_full3, id=list(n=0))
crPlots(m_full3, id=list(n=0))
# influencePlot(m_full3, id=list(n=3))
par(mfrow=c(2,2))
plot(m_full3)
par(mfrow=c(1,1))
```

The Q–Q plot reveals slight deviations in the tails, but the residuals are approximately normally distributed. A few observations appear as influential in the Cook’s distance and Residuals vs Leverage plots, which will be later assessed.

### 3) Interactions

To check for potential nonlinear relationships between predictors, several interaction models were tested. Each main variable was combined with the others. The **step()** function was used to identify significant interactions based on AIC minimization.

```{r}
md1 <- lm(sqrt(price)~bathrooms*(f.yearBuilt+levels+garageSpaces+parking+pool+homeType),data=df)
#summary(md1); step(md1)
# bathrooms:levels
# bathrooms:garageSpaces
# bathrooms:parking
# bathrooms:homeType

md2 <- lm(sqrt(price)~bedrooms*(f.yearBuilt+levels+garageSpaces+parking+pool+homeType),data=df)
#summary(md2); step(md2)
# bedrooms:f.yearBuilt
# bedrooms:levels
# bedrooms:garageSpaces
# bedrooms:pool
# bedrooms:homeType

md3 <- lm(sqrt(price)~sqrt(livingArea)*(f.yearBuilt+levels+garageSpaces+parking+pool+homeType),data=df)
# summary(md3); 
step(md3)
# sqrt(livingArea):f.yearBuilt
# sqrt(livingArea):parking
# sqrt(livingArea):homeType

md4 <- lm(sqrt(price)~f.yearBuilt*(levels+garageSpaces+parking+pool+homeType),data=df)
#summary(md4); step(md4)
# f.yearBuilt:levels
# f.yearBuilt:garageSpaces
# f.yearBuilt:parking 
# f.yearBuilt:pool
# f.yearBuilt:homeType

md5 <- lm(sqrt(price)~levels*(f.yearBuilt+garageSpaces+parking+pool+homeType),data=df)
#summary(md5); step(md5)
# levels:f.yearBuilt
# levels:garageSpaces
# levels:homeType

md6 <- lm(sqrt(price)~garageSpaces*(f.yearBuilt+levels+parking+pool+homeType),data=df)
#summary(md6); step(md6)
# garageSpaces:f.yearBuilt
# garageSpaces:levels
# garageSpaces:pool
# garageSpaces:homeType

md7 <- lm(sqrt(price)~parking*(f.yearBuilt+levels+garageSpaces+pool+homeType),data=df)
#summary(md7); step(md7)
# parking:f.yearBuilt
# parking:homeType

md8 <- lm(sqrt(price)~pool*(f.yearBuilt+levels+garageSpaces+parking+homeType),data=df)
#summary(md8); step(md8)
# pool:f.yearBuilt
# pool:garageSpaces
# pool:parking
# pool:homeType

md9 <- lm(sqrt(price)~homeType*(f.yearBuilt+levels+garageSpaces+parking+pool),data=df)
#summary(md9); step(md9)
# homeType:f.yearBuilt
# homeType:levels
# homeType:garageSpaces
# homeType:pool

model <- lm(
  sqrt(price) ~ bedrooms + bathrooms + sqrt(livingArea) + f.yearBuilt +
    levels + garageSpaces + parking + pool + homeType +
    bathrooms:levels +
    bathrooms:garageSpaces +
    bathrooms:parking +
    bathrooms:homeType +
    bedrooms:f.yearBuilt +
    bedrooms:levels +
    bedrooms:garageSpaces +
    bedrooms:pool +
    bedrooms:homeType +
    livingArea:f.yearBuilt +
    livingArea:levels +
    livingArea:garageSpaces +
    livingArea:parking +
    livingArea:homeType +
    f.yearBuilt:levels +
    f.yearBuilt:garageSpaces +
    f.yearBuilt:parking +
    f.yearBuilt:pool +
    f.yearBuilt:homeType +
    levels:garageSpaces +
    levels:homeType +
    garageSpaces:pool +
    garageSpaces:homeType +
    parking:homeType +
    pool:parking +
    pool:homeType,
  data = df
)
#summary(model)
Anova(model)
```

The final model includes several significant interaction effects, such as: bedrooms:f.yearBuilt, bedrooms:homeType, f.yearBuilt:livingArea, f.yearBuilt:levels, f.yearBuilt:garageSpaces, f.yearBuilt:parking, f.yearBuilt:pool, f.yearBuilt:homeType, garageSpaces:pool, and parking:pool.

According to the ANOVA table, these interactions were statistically significant (p < 0.05). Adding these terms improved the model’s explanatory power, increasing the Adjusted $R^2$ from 0.59 to 0.62.

## 6.2. FINAL MODEL

```{r}
model_reduced <- lm(
  sqrt(price) ~ bedrooms + bathrooms + sqrt(livingArea) + f.yearBuilt +
    levels + garageSpaces + parking + pool + homeType +
    bedrooms:f.yearBuilt +
    bedrooms:homeType +
    f.yearBuilt:sqrt(livingArea) +
    f.yearBuilt:levels +
    f.yearBuilt:garageSpaces +
    f.yearBuilt:parking +
    f.yearBuilt:pool +
    f.yearBuilt:homeType +
    garageSpaces:pool,
  data=df
)
# summary(model_reduced)
```

```{r}
marginalModelPlots(model_reduced, id=list(n=0))
residualPlots(model_reduced, id=list(n=0))
influencePlot(model_reduced, id=list(n=3))
par(mfrow=c(2,2))
plot(model_reduced)
par(mfrow=c(1,1))
```

The Q–Q plot indicates that the residuals are approximately normally distributed, although some deviations can be seen in the tails. The Scale–Location and Residuals vs Fitted plots suggest that the variance is relatively constant across fitted values, with no major heteroscedasticity detected. The Cook’s distance plot highlights a few observations with high leverage or influence, which could affect the stability of the model. Therefore, an assessment of influential data will be carried out to improve the model’s performance.

## 6.3. Influent data assesment

To assess influential observations, studentized residuals and leverage values were examined.

Observations 4222 and 4754 were identified as influential, as they presented both high residuals and high leverage values. Additionally, they were also labelled as multivariate outliers. The Cook’s distance boxplot confirmed that one of these points was notably distant from the rest. Therefore, these observations were removed to evaluate whether their exclusion improves the model’s stability and performance.

```{r}
llres <- which(abs(rstudent(model_reduced)) > 3)
llhat <- which(hatvalues(model_reduced) > 3*length(coef(model_reduced))/nrow(df))
common_outliers <- intersect(llres, llhat); common_outliers
df[common_outliers, ]
llcoo <- Boxplot(cooks.distance(model_reduced), id=list(n=10,labels=row.names(df)))
df <- df[-common_outliers,] # remove influent data
```

We fit the model again without the 2 outliers.

```{r}
model_reduced_2 <- lm(
  sqrt(price) ~ bedrooms + bathrooms + sqrt(livingArea) + f.yearBuilt +
    levels + garageSpaces + parking + pool + homeType +
    bedrooms:f.yearBuilt +
    bedrooms:homeType +
    f.yearBuilt:livingArea +
    f.yearBuilt:levels +
    f.yearBuilt:garageSpaces +
    f.yearBuilt:parking +
    f.yearBuilt:pool +
    f.yearBuilt:homeType +
    garageSpaces:pool,
  data=df
)

summary(model_reduced_2)
```

We compute the influential data again to check if any influential observations remain. No additional points were detected, confirming that the model is now stable and not affected by outliers.

```{r}
llres <- which(abs(rstudent(model_reduced_2)) > 3)
llhat <- which(hatvalues(model_reduced_2) > 3 * length(coef(model_reduced_2)) / nrow(df))
common_outliers <- intersect(llres, llhat); common_outliers
llcoo <- Boxplot(cooks.distance(model_reduced_2), id=list(n=10,labels=row.names(df)))
```

After removing 2 influential data, our model achieves a $R^2=0.616$.

## 6.4. Model validation
The final model is validated through diagnostic plots after removing the influential data. The Marginal Model Plots show an improvement compared to the initial version. In particular, the bedrooms relationships now align more closely with the model line, indicating that the fitted values capture the observed trend more accurately. The relationship between sqrt(livingArea) and the target remains well described, with the model line closely following the data pattern.

The Residuals vs Fitted plot exhibits a random distribution of residuals around zero, suggesting that the linearity and homoscedasticity assumptions hold. The Scale–Location plot confirms this, as the spread of residuals is relatively constant across fitted values. The Q–Q plot shows a deviation at the tails, but overall the residuals are approximately normally distributed.

The Residuals vs Leverage and Cook’s Distance plots indicate that no additional influential points remain after removing observations 4222 and 4754.

```{r}
marginalModelPlots(model_reduced_2, id=list(n=0))
avPlots(model_reduced_2, id=list(n=0))
influencePlot(model_reduced_2, id=list(n=3))
par(mfrow=c(2,2))
plot(model_reduced_2)
par(mfrow=c(1,1))
```
