---
title: "Model_SIM.Rmd"
output:
  pdf_document: default
  html_document: default
date: "2025-10-24"
---
```{r}
# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())
```

```{r}
df <- read.csv("House_Clean.csv", header = TRUE)

df[sapply(df, is.character)] <- lapply(df[sapply(df, is.character)], as.factor)

# Converting binary columns (only 0 or 1) to factor
df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))] <- 
  lapply(df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))], as.factor)

summary(df)
```

```{r}
num_vars <- names(df)[sapply(df, is.numeric)]
cor(df[,num_vars]) 
plot(df[,num_vars])
```
## Modeling using explanatory numeric variables

How to define the best model? Iterative process.
Suggestion: after data preparation etc. 
    Modelling
  1) Use numeric variables.
Try to find the best transformation needed to the target variable or the explanatory variables (improve the linear )
  2) Add factors (only the main effects)
  3) Add interactions
  
MLR has explanatory variables with only numeric 
When I add factors it becomes GLM 

# Model: numeric variables
```{r}
head(df)
m1 <- lm( price ~., data=df[,num_vars])
summary(m1)
# mirando p values: bedrooms not sig. yearbuilt??
# reminder: intercept beta1. beta4 ~ 0

Anova(m1)
model.matrix(m1) #useless for practical exercise but this is the mode matrix (Y = X*betas)
```

## Transformations

```{r}
summary(m2)
boxcox(prestige ~ education + income, data = df[, c(1:4)])
boxTidwell(prestige ~ education + income, data = df[, c(1:4)]) #income needed
m3 <- lm( prestige ~ education + log(income), data = df[, c(1:4)])
summary(m3)
par(mfrow=c(2,2))
plot(m3, id.n=0)
par(mfrow=c(1,1))

marginalModelPlots( m3, id=list(n=0))
residualPlots( m3, id=list(n=0))
avPlots( m3, id=list(n=0))
crPlots( m3, id=list(n=0))
influencePlot( m3, id=list(n=3))
```


# Anova - Net-effects of explanatory variables
(Es cuando metes los factors al modelo)

Response: prestige
          Sum Sq Df  F value    Pr(>F)    
education 7142.4  1 116.0109 < 2.2e-16 ***
income    1376.7  1  22.3612 7.579e-06 ***
women        5.3  1   0.0858    0.7702    
Residuals 6033.6 98                       

Es un p-value condicional:
- once income + women are already in the model, adding education is... (pvalue~0: significative)
- once income + education are already in the model, adding women is... (pvalue>>0: not significative) (useless)

  - cuidado: anova != Anova

```{r}
# step is not based on inferential (tradeoff goodness of fit/complexity)
m2 <- step( m1 )
```
('achaiche'??)
comparas si quitas women, si no quitas ninugno, si quitas income, si quitas education...

# Start:  AIC=424.17
prestige ~ education + income + women

            Df Sum of Sq     RSS    AIC
- women      1       5.3  6038.9 422.26
<none>                    6033.6 424.17
- income     1    1376.7  7410.3 443.14
- education  1    7142.4 13176.0 501.84

Step:  AIC=422.26
prestige ~ education + income

            Df Sum of Sq     RSS    AIC
<none>                    6038.9 422.26
- income     1    2248.1  8287.0 452.54
- education  1    8577.3 14616.2 510.42

BIC-Schwartz

```{r}
summary(m2)
par(mfrow=c(2,2))
plot(m2, id.n=0)

# residual analysis looks fine. normal distribution tmb

# units for resiudals: Y-^Y
# standardized residuals: should be interpreted as standard normal distribution. very convenient (simlpifies interpretation)

par(mfrow=c(1,1))
influencePlot( m2, id=list(n=5,method="noteworthy") )

# Studentized Residuals (just a technichality)
# transformed residual interpreted as N~(0,1)
# vertical lines: threshold (la primera: 2xmean head value, la segunda 3xmhv)
# horizontal lines: unusual pero no tanto como general managers x ej

residualPlots(m2) #useful for assignment
# checkeas los residuals y los explanatory variables a ver si no hay alguna con tendencia
# income parece un poco raro... we have to try to improve linear relation bw prestige and income
```
 residualPlots(m2)
           Test stat Pr(>|Test stat|)    
education     1.3447        0.1818236    
income       -3.6732        0.0003904 ***
Tukey test   -1.5264        0.1269206 
-- te avisa que en income pasa algo

```{r}
marginalModelPlots(m2)
# if you see discrepancies your model has to be improved
avPlots(m2)
```

