---
title: "Model_SIM.Rmd"
output:
  pdf_document: default
  html_document: default
date: "2025-10-24"
---
```{r}
# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())

library(car)
library(FactoMineR)
library(lmtest)
library(effects)
library(AER)
library(MASS)
```

```{r}
df <- read.csv("House_Clean2.csv", header = TRUE)

df[sapply(df, is.character)] <- lapply(df[sapply(df, is.character)], as.factor)

# Converting binary columns (only 0 or 1) to factor
df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))] <- 
  lapply(df[sapply(df, function(x) all(na.omit(unique(x)) %in% c(0, 1)))], as.factor)

summary(df)
```

```{r}
num_vars <- names(df)[sapply(df, is.numeric)]
cor(df[,num_vars]) 
plot(df[,num_vars])
```
## Modeling using explanatory numeric variables

How to define the best model? Iterative process.
Suggestion: after data preparation etc. 
    Modelling
  1) Use numeric variables.
Try to find the best transformation needed to the target variable or the explanatory variables (improve the linear )
  2) Add factors (only the main effects)
  3) Add interactions
  
MLR has explanatory variables with only numeric 
When I add factors it becomes GLM 

# Model: numeric variables
We build a model with all numeric variables. As livingArea and pricePerSquareFoot perfectly describe the target, we use pricePerSquareFoot because after building both models we get better results for the model using pricePerSquareFoot.

Analyzing the variance inflation factor (VIF), we observe that all values are below 5, indicating that the predictors are not highly linearly correlated, so multicollinearity is not a problem.

```{r}
head(df)
m0 <- lm(price ~bedrooms+bathrooms+livingArea+n.yearBuilt, data=df)
summary(m0)
vif(m0)

m0_f <- lm(price ~bedrooms+bathrooms+livingArea+f.yearBuilt, data=df)
summary(m0_f)
vif(m0_f)
# both models have similar AIC and R^2, so we keep the factor yearsBuilt

```
We use boxcox transformation and we observe that lambda is near 0.5, so we apply sqrt transformation to our target price.

We compare two models m1 and m2, which are the models with sqrt of price and the other variables, with and without n.yearsBuilt. In the anova test, p-value < 0.05, so we reject the null hypothesis, meaning that the big model is better, so we keep n.yearBuilt. We also check the AIC value, and this is confirmed because AIC for the model 2 is bigger than AIC for the model 1.

bedrooms lambda negatiu, pr < 0.05 -> reject H0 -> relació no es lineal -> transformació -> no transformació
bathrooms lambda positiu, pr < 0.05 -> reject H0 -> relació no es lineal -> transformació -> sqrt
livingArea lambda positiu, pr < 0.05 -> reject H0 -> relació no es lineal -> transformació -> 1/3


```{r}
boxcox(price ~bedrooms+bathrooms+livingArea+n.yearBuilt, data=df) #we see that sqrt is the suitable tranformation
boxTidwell(sqrt(price) ~ bedrooms+bathrooms+livingArea+n.yearBuilt, data=df) #we put sqrt to livingArea

m1 <- lm(sqrt(price) ~bedrooms+bathrooms+livingArea+n.yearBuilt, data=df)
summary(m1)
vif(m1)
step(m1)
AIC(m1)
par(mfrow=c(2,2))
plot(m1,id=list(n=Inf,labels=row.names(df)))
par(mfrow=c(1,1))

#we remove n.yearBuilt -> higher AIC
m2 <- lm(sqrt(price) ~ bedrooms+bathrooms+livingArea, data=df)
summary(m2)
vif(m2)
anova(m2,m1)
step(m2)
AIC(m2)

#we use the sqrt to livingArea transformation
m3_n <- lm(sqrt(price) ~bedrooms+bathrooms+sqrt(livingArea)+n.yearBuilt, data=df)
summary(m3_n)
vif(m3_n)
anova(m3_n,m1)
step(m3_n)
AIC(m3_n)

#we try to use the factor version of yearBuilt and it yields better results, so we keep this model
m3 <- lm(sqrt(price) ~bedrooms+bathrooms+sqrt(livingArea)+f.yearBuilt, data=df)
summary(m3)
vif(m3)
anova(m3,m1)
step(m3)
AIC(m3)
```
*We keep m3 as the numeric model*. We obtain for it a flatter line in Residuals vs Fitted than with model 1. This means the transformation I(livingArea^(1/3)) corrected the non-linearity that existed in m1.

AIC is also lower 67552.73 < 67707.

QQ plot: There’s a small deviation in the tails (some high-value houses still don’t fit perfectly), but the central part follows the line closely.

Looking at the crPlot, a component + residual plot, it gives information of how well each numeric predictor fits a linear pattern after accounting for the others.

```{r}
marginalModelPlots(m3, id=list(n=0))
residualPlots(m3, id=list(n=0))
avPlots(m3, id=list(n=0))
crPlots(m3, id=list(n=0))
influencePlot(m3, id=list(n=3))
plot(m3)
plot(allEffects(m3))
```


# Model with bathrooms and bedrooms as factors
We try to use bathrooms and bedrooms as factors, obtaining Rsquared of 0.5725 for numeric and Rsquared of 0.5542 for factors, and AIC:
df      AIC
m3    6 67529.17
mf_2  8 67762.35
So we use the numeric version of the variables.


### Starting from the model with all variables 

```{r}
# Model with all factors
m_full <- lm(sqrt(price) ~bedrooms+bathrooms+sqrt(livingArea)+f.yearBuilt+levels+garageSpaces+hasGarage+spa+parking+pool+homeType, data=df)
summary(m_full)
Anova(m_full)

# We see that pvalues for hasGarage and spa are lower, we try to exclude them from the model

# Model without hasGarage
m_full2 <- lm(sqrt(price) ~bedrooms+bathrooms+sqrt(livingArea)+f.yearBuilt+levels+garageSpaces+spa+parking+pool+homeType, data=df)
summary(m_full2)
Anova(m_full2)

# Model without hasGarage and spa
m_full3 <- lm(sqrt(price) ~bedrooms+bathrooms+sqrt(livingArea)+f.yearBuilt+levels+garageSpaces+parking+pool+homeType, data=df)
summary(m_full3)
Anova(m_full3)

# Model without spa
m_full4 <- lm(sqrt(price) ~bedrooms+bathrooms+sqrt(livingArea)+f.yearBuilt+levels+garageSpaces+hasGarage+parking+pool+homeType, data=df)
summary(m_full4)
Anova(m_full4)

# AICs are similar, so we keep the simplest model
        #df      AIC
#m_full  20 67250.86
#m_full2 19 67257.26
#m_full3 18 67267.14
#m_full4 19 67260.52

m_step <- step(m_full3, direction = "both", trace = TRUE)
summary(m_step)
AIC(m_step)

marginalModelPlots(m_full3, id=list(n=0))
residualPlots(m_full3, id=list(n=0))
avPlots(m_full3, id=list(n=0))
crPlots(m_full3, id=list(n=0))
influencePlot(m_full3, id=list(n=3))
par(mfrow=c(2,2))
plot(m_full3)
par(mfrow=c(1,1))
```
### Interactions

```{r}
md1 <- lm(sqrt(price)~bathrooms*(f.yearBuilt+levels+garageSpaces+parking+pool+homeType),data=df)
# summary(md1)
step(md1)
# with the step function we conclude that the meaningful interactions with bathroom are
# bathrooms:levels
# bathrooms:garageSpaces
# bathrooms:parking
# bathrooms:homeType

md2 <- lm(sqrt(price)~bedrooms*(f.yearBuilt+levels+garageSpaces+parking+pool+homeType),data=df)
# summary(md2)
step(md2)
# bedrooms:f.yearBuilt
# bedrooms:levels
# bedrooms:garageSpaces
# bedrooms:pool
# bedrooms:homeType

md3 <- lm(sqrt(price)~livingArea*(f.yearBuilt+levels+garageSpaces+parking+pool+homeType),data=df)
# summary(md3)
step(md3)
# livingArea:f.yearBuilt
# livingArea:levels
# livingArea:garageSpaces
# livingArea:parking
# livingArea:homeType

md4 <- lm(sqrt(price)~f.yearBuilt*(levels+garageSpaces+parking+pool+homeType),data=df)
# summary(md4)
step(md4)

#f.yearBuilt:levels
# f.yearBuilt:garageSpaces
# f.yearBuilt:parking 
# f.yearBuilt:pool
# f.yearBuilt:homeType

md5 <- lm(sqrt(price)~levels*(f.yearBuilt+garageSpaces+parking+pool+homeType),data=df)
# summary(md5)
step(md5)
# levels:f.yearBuilt
# levels:garageSpaces
# levels:homeType

md6 <- lm(sqrt(price)~garageSpaces*(f.yearBuilt+levels+parking+pool+homeType),data=df)
# summary(md6)
step(md6)
# garageSpaces:f.yearBuilt
# garageSpaces:levels
# garageSpaces:pool
# garageSpaces:homeType

md7 <- lm(sqrt(price)~parking*(f.yearBuilt+levels+garageSpaces+pool+homeType),data=df)
# summary(md7)
step(md7)
# parking:f.yearBuilt
# parking:homeType

md8 <- lm(sqrt(price)~pool*(f.yearBuilt+levels+garageSpaces+parking+homeType),data=df)
# summary(md8)
step(md8)
# pool:f.yearBuilt
# pool:garageSpaces
# pool:parking
# pool:homeType

md9 <- lm(sqrt(price)~homeType*(f.yearBuilt+levels+garageSpaces+parking+pool),data=df)
# summary(md9)
step(md9)
# homeType:f.yearBuilt
# homeType:levels
# homeType:garageSpaces
# homeType:pool

model <- lm(
  sqrt(price) ~ bedrooms + bathrooms + sqrt(livingArea) + f.yearBuilt +
    levels + garageSpaces + parking + pool + homeType +
    bathrooms:levels +
    bathrooms:garageSpaces +
    bathrooms:parking +
    bathrooms:homeType +
    bedrooms:f.yearBuilt +
    bedrooms:levels +
    bedrooms:garageSpaces +
    bedrooms:pool +
    bedrooms:homeType +
    livingArea:f.yearBuilt +
    livingArea:levels +
    livingArea:garageSpaces +
    livingArea:parking +
    livingArea:homeType +
    f.yearBuilt:levels +
    f.yearBuilt:garageSpaces +
    f.yearBuilt:parking +
    f.yearBuilt:pool +
    f.yearBuilt:homeType +
    levels:garageSpaces +
    levels:homeType +
    garageSpaces:pool +
    garageSpaces:homeType +
    parking:homeType +
    pool:parking +
    pool:homeType,
  data = df
)

#summary(model)
# Anova(model)
```
### FINAL MODEL

```{r}

model_reduced <- lm(
  sqrt(price) ~ bedrooms + bathrooms + sqrt(livingArea) + f.yearBuilt +
    levels + garageSpaces + parking + pool + homeType +
    bedrooms:f.yearBuilt +
    bedrooms:homeType +
    f.yearBuilt:livingArea +
    f.yearBuilt:levels +
    f.yearBuilt:garageSpaces +
    f.yearBuilt:parking +
    f.yearBuilt:pool +
    f.yearBuilt:homeType +
    garageSpaces:pool,
  data=df
)

#summary(model_reduced)
```

```{r}
marginalModelPlots(model_reduced, id=list(n=0))
#residualPlots(model_reduced, id=list(n=0))
avPlots(model_reduced, id=list(n=0))
#crPlots(model_reduced, id=list(n=0))
influencePlot(model_reduced, id=list(n=3))
par(mfrow=c(2,2))
plot(model_reduced)
par(mfrow=c(1,1))
```

### Influent data assesment

```{r}
llres<-which(abs(rstudent(model_reduced))>3); llres

llhat<-which(hatvalues(model_reduced)>3*length(coef(model_reduced))/nrow(df));llhat

llcoo<-Boxplot(cooks.distance(model_reduced), id=list(n=10,labels=row.names(df)))
abline(h=3*length(coef(model_reduced))/nrow(df), col ='red')

influencePlot(model_reduced)
df[llres,]
ll<-which(row.names(df) %in% c("611", "1300"));ll

#df<-df[-ll,]
```

## Influent data but only the common observations in llres and llhat
# it's only one -> 611 we can remove it
```{r}
llres <- which(abs(rstudent(model_reduced)) > 3)
llhat <- which(hatvalues(model_reduced) > 3 * length(coef(model_reduced)) / nrow(df))
common_outliers <- intersect(llres, llhat)
df[common_outliers, ]

df <- df[-common_outliers, ]
```

